\chapter{Introduction}\label{chap:intro}

This research,
\textit{RicoSLAM: localization and mapping in dynamic environments},
aims to advance the current state of the art in \gls{slam}.
This work focuses on improving localization and mapping performance when
performing \gls{slam} in dynamic environments, using wheeled odometry and 2D~laser
scanner data from sensors onboard industrial ground mobile robots.
\gls{slam} in dynamic environments is closely tied to lifelong~\gls{slam} and
long-term operation of mobile robots, enabling robots to maintain accurate
localization and mapping over extended periods while adapting to continuous
environmental changes.

This chapter is divided into seven sections.
First, \secref{sec:intro:context} contextualizes this research work and
describes the primary challenges related to localization and mapping in dynamic
environments.
\secref{sec:intro:motivation} outlines the key motivations and reasons for
undertaking this research.
Next, \secref{sec:intro:problem} introduces the problem that this PhD~Thesis
seeks to address.
\secref{sec:intro:rq+h} presents the main research questions and hypotheses that
guided the research work.
Then, \secref{sec:intro:scope} defines the scope and objectives of the
PhD~Thesis.
\secref{sec:intro:contributions} highlights the unique contributions the
research makes to the literature, emphasizing its novelty and impact.
Lastly, \secref{sec:intro:structure} outlines the document's organization.
A final note is provided regarding the funding that supported the work carried
out throughout the PhD~Thesis.

\section{Context}\label{sec:intro:context}
Nowadays, \glspl{amr} are being increasingly deployed in
several industries, including logistics, retail, warehousing, automotive and
aerospace manufacturing, agriculture, and healthcare.
\glspl{amr} may automate tasks such as material transport, automated inspection,
and security monitoring.
By improving operational efficiency, flexibility and productivity,
\glspl{amr} help companies achieve higher efficiencies and
reduce production costs.
Consequently, these improvements contribute to the growth of the global economy
and increase the worldwide interest in the \gls{amr} market.
According to a study by ABI Research, the global robotics market is estimated to
be roughly US\$50 billion in 2025, growing at a \gls{cagr}
of about 14\% to reach approximately US\$111 billion by 2030.
This market growth is driven by labour workforce shortages, supply chain
reformation, and high turnover rates, with the concurrent increasing consumer
demand. Mobile robots in particular can mitigate these challenges by expediting
the loading/unloading of trucks, automating inventory management, and performing
last-mile delivery services~\cite{abi-research:tech-report:2025}.

An \gls{amr} must interpret its surroundings by extracting meaningful data from
its sensors to localize itself within the environment, enabling autonomous
navigation~\cite{siegwart:book:2004}.
\gls{slam} addresses this perception and localization challenge by concurrently
incorporating the robot state estimation into the mapping process to build a
representation of the environment as perceived by the robot.
The state is typically described by the robot's pose (position and orientation),
although other variables, such as velocity and calibration parameters, may also
be included in the state. As for the environment representation, the mapping
process involves creating a map incrementally, built as the robot explores
unknown areas or refines it through passages in known locations, to describe
the environment in which the robot operates. The map representation serves
multiple purposes, such as path planning, providing a visual reference for
human operators, while allowing the robot to reset its localization error when
revisiting known areas~\cite{cadena:tro:2016}.

In a static scene, the robot would only need to map once because the map and
sensor data would remain consistent with the environment throughout time.
However, industrial and intralogistics locations, agricultural sites,
warehouses, and service-oriented applications, such as shopping centers and
healthcare institutions, have inherently moving elements in the scene
(humans, objects), reconfiguration of the environment
(reorganization of machines, patient beds), and appearance variations
(lighting, weather, seasonal, or daytime changes).
The deployment of \glspl{amr} in these varying conditions are a challenge for
the \gls{slam} system, where the system should decide how (consider the most
current state, the evolution of the environment dynamics, or only the most
static changes?) and when to update the map (when the variations occur, or only
after a certain time?). This challenge is also known as the stability-plasticity
dilemma, where a lifelong~\gls{slam} system should both adapt to new environment
changes and preserve old states over time~\cite{biber-duckett:ijrr:2009}.
Furthermore, the map would grow indefinitely when gathering new information from
the environment without any map management system. This ever-growing problem
poses another challenge for the \gls{slam} system in the long-term due to the
limited computational resources of the mapping vehicle. Indeed, the map size
should be dependent on the environment area and not on the trajectory length of
the mapping process nor operation time, only growing when the robot would
explore unknown locations~\cite{kretzschmar-stachniss:ijrr:2012}.

\todo{Brief literature analysis (only after finishing Literature Review chapter)}

\section{Motivation}\label{sec:intro:motivation}
\section{Problem Statement}\label{sec:intro:problem}
\section{Research Questions and Hypotheses}\label{sec:intro:rq+h}
\section{Scope and Objectives}\label{sec:intro:scope}

\section{Contributions}\label{sec:intro:contributions}
TBC

\subsection{Main contributions}
\begin{description}[leftmargin=1cm,style=nextline,font=\normalfont]
\item[\cite{sousa:jfr:2023}] \fullcite{sousa:jfr:2023}
\item[\cite{sousa:icara:2024}] \fullcite{sousa:icara:2024}
\item[\cite{sousa:icarsc:2025}] \fullcite{sousa:icarsc:2025}
\end{description}
\subsection{Complementary contributions}
\begin{description}[leftmargin=1cm,style=nextline,font=\normalfont]
\item[\cite{ribeiro:submitted:2025}] \fullcite{ribeiro:submitted:2025}
\item[\cite{lopes:submitted:2026}] \fullcite{lopes:submitted:2026}
\end{description}
\subsection{Public datasets}
\begin{description}[leftmargin=1cm,style=nextline,font=\normalfont]
\item[\cite{ribeiro:dataset:2025}] \fullcite{ribeiro:dataset:2025}
\end{description}

\section{Document Structure}\label{sec:intro:structure}

\section*{Funding}
This work has received funding from National Funds through the Portuguese
funding agency, FCT -- Fundação para a Ciência e a Tecnologia, within
scholarship 2021.04591.BD
(DOI: \href{https://doi.org/10.54499/2021.04591.BD}{10.54499/2021.04591.BD})
and project LA/P/0063/2020
(DOI: \href{https://doi.org/10.54499/LA/P/0063/2020}{10.54499/LA/P/0063/2020}).
This work has also been co-financed by Component 5 -- Capitalisation
and Business Innovation, integrated in the Resilience Dimension of the Recovery
and Resilience Plan within the scope of the Recovery and Resilience
Mechanism~(MRR) of the European Union (EU), framed in the Next Generation~EU,
for the period 2021--2026, within project GreenAuto, with reference 54.
