\chapter{Conclusions and Future Work}\label{chap:conclusions}

This PhD~Thesis, \emph{RicoSLAM: localization and mapping in dynamic
environments}, researched how to perform \gls{slam} with a ground mobile
robot operating in dynamic indoor environments, using wheeled odometry and
2D~laser scanner data as the only sensing modalities. The research was
organized around the development of a corner-like feature detector
(\chapref{chap:features}), a 2D~point cloud registration framework
based on distance maps (\chapref{chap:distance-maps}), and the
RicoSLAM \gls{slam} framework itself (\chapref{chap:slam}), with the
two research questions and the corresponding hypotheses formulated in
\secref{sec:intro:rq+h} as the underlying motivation.
This chapter consolidates the main conclusions drawn from
these contributions and identifies the most relevant directions
for future research.

This chapter is divided into two sections.
First, \secref{sec:conclusions:contributions} summarizes the
conclusions and contributions of the PhD~Thesis.
Then, \secref{sec:conclusions:final-remarks} presents final remarks and
the directions of future work, including the involvement of
the candidate in the scientific community throughout the PhD~Thesis.

\section{Conclusions and Contributions}\label{sec:conclusions:contributions}

The research questions and hypotheses formulated in
\secref{sec:intro:rq+h} guided the development of the contributions
reported in \chapref{chap:features} to \chapref{chap:slam}.
On the localization side, the corner-like
feature detector of \chapref{chap:features}, the distance map-based
registration framework of \chapref{chap:distance-maps}, and the
local-map operating mode of RicoSLAM (\chapref{chap:slam}) address
the hypotheses on geometric features
(Hypothesis~\ref{hypothesis:localization:features}), distance maps
as a correspondence-free representation
(Hypothesis~\ref{hypothesis:localization:distance-maps}), and local
maps continuously updated with the latest sensor observations
(Hypothesis~\ref{hypothesis:localization:local-maps}).
On the mapping side, the online and offline dynamics removal mechanisms
of \secref{sec:slam:dynamics}, the relocalization policy of
\secref{sec:slam:online:relocalization}, and the distance map-based \gls{gba}
formulations proposed in \secref{sec:slam:gba} cover the hypotheses on ray
tracing-based map updating (Hypothesis~\ref{hypothesis:mapping:ray-tracing})
and cleaner static maps via offline dynamics removal
(Hypothesis~\ref{hypothesis:mapping:clean-map}),
bounded graph growth through relocalization
(Hypothesis~\ref{hypothesis:mapping:relocalization}), and offline global
map refinement (Hypothesis~\ref{hypothesis:mapping:ogba}).
The remainder of this section summarizes the main conclusions drawn from
the corresponding chapters.

\chapref{chap:features} introduces a line fitting-based corner-like
detector for 2D~laser scanner data~\cite{sousa:icara:2024}, in which
the keypoint is computed as the closed-form intersection of two
locally fitted lines at multiple distance-based neighborhood scales. This
formulation avoids the dependency on global line-segment extractors
that characterizes most literature corner detectors and produces
feature locations that do not need to coincide with a particular
laser return. Experimental results showed that the proposed
corner-like features exhibit viewpoint invariance, where
$6$~features attained a $100\%$ observability ratio under the
considered sensor \gls{fov} and pose uncertainty, while $20$ and $16$ features
remained observable above $50\%$ and $75\%$ across the experiments, respectively.
In addition, the analysis of the association distance among the $39$
most persistent features showed that $91.5\%$ of the possible
feature associations have a standard deviation at or below
$\SI{0.03}{\metre}$, consistent with the characteristic noise of the
Hokuyo~UST-10LX 2D~laser scanner sensor while yielding geometric invariance.
These results indicate that the proposed corner-like features are stable enough
to be used as input to a feature-based tracking module for mobile robots.

\chapref{chap:distance-maps} introduces a 2D~point cloud registration
framework that replaces the explicit nearest-neighbor search performed
at every iteration of classical
\gls{icp}~\cite{besl-mckay:tpami:1992,chen-medioni:icra:1991} with up
to $O(1)$ grid lookups on a precomputed distance map data structure.
Both point-to-point and point-to-plane error formulations are derived
within the Gauss-Newton framework on the $SE(2)$ manifold, with the
Jacobians of both variants obtained through right-sided perturbation.
The unified distance map data structure introduced in this chapter
supports dense and sparse storage layouts and three representation
levels~$\mathcal{M}_l$, while remaining compatible with the classical
\gls{icp} formulations used as baselines. Across the synthetic and
IILABS~3D~\cite{ribeiro:access:2025} benchmarks, the precomputed
point-to-point variant $\mathcal{D}_\mathrm{p2p}(\mathcal{M}_1)$ stood
out as the best-performing point-to-point formulation, consistently
improving over its analytical counterparts
$\mathcal{D}_\mathrm{p2p}(\mathcal{M}_0)$ and
$\mathrm{ICP}_{\mathcal{D},\,\mathrm{p2p}}$ on the \gls{rte} and
\gls{rre} metrics across all four planar sequences of the
IILABS~3D~\cite{ribeiro:access:2025} dataset (Nav~A~Diff, Nav~A~Omni,
Loop, and Slippage), in agreement with the synthetic results
discussed in \chapref{chap:distance-maps}. More importantly,
$\mathcal{D}_\mathrm{p2p}(\mathcal{M}_1)$ also reached competitive
accuracy against the analytical point-to-plane formulations
$\mathcal{D}_\mathrm{p2pl}(\mathcal{M}_0)$ and
$\mathrm{ICP}_{\mathcal{D},\,\mathrm{p2pl}}$. On Nav~A~Omni
at $r_\mathrm{max} = \SI{30}{\metre}$, $\mathcal{D}_\mathrm{p2p}(\mathcal{M}_1)$
obtained the lowest drift overall both on \gls{rte}
(\SI{0.81}{\percent}) and on \gls{rre} (\SI{0.063}{\degree\per\metre}).
On Nav~A~Diff at the same range, $\mathcal{D}_\mathrm{p2p}(\mathcal{M}_1)$
attained the lowest \gls{rre} overall (\SI{0.078}{\degree\per\metre}) with
a \gls{rte} of \SI{0.95}{\percent}, within \SI{0.07}{\percent} of the
$\mathrm{ICP}_{\mathcal{D},\,\mathrm{p2pl}}$ baseline. On the Loop sequence,
the point-to-plane formulations retained an advantage, with
$\mathrm{ICP}_{\mathcal{D},\,\mathrm{p2pl}}$ attaining
\SI{1.37}{\percent} \gls{rte} and \SI{0.106}{\degree\per\metre}
\gls{rre} against \SI{2.14}{\percent} and \SI{0.123}{\degree\per\metre}
for $\mathcal{D}_\mathrm{p2p}(\mathcal{M}_1)$, consistent with the
chapter's discussion of the residual gap that emerges when traversing
large spaces through narrow passages. Overall, these results
underline the relevance of the precomputed distance field to
regularize point-to-point correspondences against sensor noise and to
provide an effective alternative to analytical point-to-plane
formulations in a broad range of operating conditions,
avoiding kd-tree search~\cite{bentley:cacm:1975} traditionally required at every
\gls{icp}~\cite{besl-mckay:tpami:1992,chen-medioni:icra:1991} iteration.

\chapref{chap:slam} presents the RicoSLAM framework developed within the scope
of this PhD~Thesis, including the distance map-based tracker from
\chapref{chap:distance-maps} with split and merge triggering criteria for
keyframe management, a loop closure detection procedure based on world-frame
proximity and graph-distance, and a relocalization policy that
transfers the tracker into an existing keyframe of the back-end graph
whenever a short-loop candidate is available. On the
IILABS~3D~\cite{ribeiro:access:2025} dataset, RicoSLAM attains the
lowest \gls{ate} across its operating modes on all four planar
sequences considered in the trajectory accuracy benchmark, with
\SI{0.047}{\metre} on Nav~A~Diff (\emph{local} mode),
\SI{0.030}{\metre} on Nav~A~Omni (\emph{local~w/upd.}\ mode),
\SI{0.033}{\metre} on Loop (\emph{local~w/upd.}\ mode), and
\SI{0.014}{\metre} on Slippage (\emph{scan~w/reloc.}\ mode). These
values are lower than the \SI{0.053}{\metre}, \SI{0.061}{\metre},
\SI{0.040}{\metre}, and \SI{0.042}{\metre} obtained by the best
SLAM~Toolbox~\cite{macenski-jambrecic:joss:2021} configurations, and
substantially lower than the GMapping~\cite{grisetti:tro:2007}
results, which fails to close the loop on the Loop sequence.
The offline pose-correction stream proposed in
\secref{sec:slam:online:pose-history} further reduces the \gls{ate} on
the three longer sequences, while its dispersion analysis suggests
that the residual error is dominated by systematic sources
(\eg, extrinsic calibration, sensor noise, ground-truth precision)
rather than by front-end drift or loop closure misalignment. In
terms of scalability, all RicoSLAM variants produce smaller back-end graphs than
SLAM~Toolbox~\cite{macenski-jambrecic:joss:2021} for comparable, and
often better, trajectory accuracy. The combination of the
relocalization policy with the local-map operating mode
(\emph{local~w/reloc.}\ variant) produces a back-end whose size
depends primarily on the spatial extent of the explored environment
rather than on the duration of operation or the path length, even
when revisiting the same locations multiple times.

Furthermore, the dynamics removal strategy proposed in \chapref{chap:slam}
operates at two complementary stages of RicoSLAM. The online stage
runs during local map maintenance and treats each laser beam as an
observation from the surrounding scene that may either see through a
previously mapped point or add new observations to the local
representation. The offline stage runs during 2D~occupancy grid map
generation and combines the same ray model with a
Bresenham-like~\cite{bresenham:ibm:1965,zingl:uas:2016} line
rasterization to suppress dynamic returns before integrating the
scans into the final occupancy grid. The experimental evaluation
considered the IILABS~3D~\cite{ribeiro:access:2025} dataset together
with four additional datasets from real-world \gls{amr} deployments
(Avantgarde, Cappero, Flowbotic, and Kuka), spanning sensor sampling
rates from \SI{6}{\hertz} to \SI{40}{\hertz}, sequence durations
from a few minutes up to roughly \SI{78}{\minute}, and path lengths
from tens of meters up to kilometers. Across every sequence,
operating mode, and threshold considered in the experiment, the
proposed offline dynamics removal mechanism consistently sharpened
the empirical occupancy distribution of the final probability map
relative to the standard rasterization baseline and reclassified to
an \emph{unknown} state cells whose only supporting observations were
transient dynamic returns. Together with the trajectory accuracy
and scalability results, these mapping experiments demonstrate that
RicoSLAM is able to produce a cleaner and more compact representation
of the environment suitable for downstream localization in
autonomous operation, even in the presence of dynamic returns and
sensor noise from industrial \gls{amr} deployments.

Lastly, the distance map-based \gls{gba} factors introduced in
\secref{sec:slam:gba} augment the pose-pose constraints of the online
graph with per-point distance map factors, both in their
point-to-point and point-to-plane formulations. The Jacobians of
these factors are derived via right-sided perturbation on the
$SE(2)$ manifold, building on the corresponding error formulations
from \chapref{chap:distance-maps}. The synthetic benchmark evaluated
four \gls{gba} factor configurations on a Manhattan-like grid of
$289$ poses perturbed by $\pm\SI{1.5}{\metre}$ on translation and
$\pm\SI{20}{\degree}$ on orientation, considering noise-free
conditions and a Gaussian range noise of
$\sigma = \SI{0.03}{\metre}$. Although the graph topology and the
synthetic scene are regular and well-structured, the results
indicate that the proposed factors are suited to globally refining
the back-end pose graph produced by the online RicoSLAM pipeline.
Consistent with the registration evaluation of
\chapref{chap:distance-maps}, the precomputed point-to-point variant
$\mathcal{D}_\mathrm{p2p}(\mathcal{M}_1)$ with voxelization of the
moving point clouds attained the lowest translational and rotational
errors under range noise conditions, reinforcing the relevance of
distance map-based point-to-point formulations even when compared
against point-to-plane variants. The full integration of the
proposed \gls{gba} stage with the online RicoSLAM pipeline and its
end-to-end validation in real-world environments remains a work in
progress at the time of writing this PhD~Thesis, and is identified as
the main open direction in \secref{sec:conclusions:final-remarks}.

Beyond the scientific contributions discussed above,
the PhD~Thesis materializes into a set of peer-reviewed publications
covering the systematic literature review on long-term localization
and mapping~\cite{sousa:jfr:2023} (\appxref{appx:slr}), the corner-like feature
detector~\cite{sousa:icara:2024} (\appxref{appx:features}), the multimodal
mobile platform~\cite{sousa:icarsc:2025} (\appxref{appx:hangfaq2})
used for data acquisition of the IILABS~3D~\cite{ribeiro:access:2025}
(\appxref{appx:iilabs3d}), and the usage of RicoSLAM during container unloading
applications~\cite{lopes:icara:2026} (\appxref{appx:cargo}), complemented by
the public release of the IILABS~3D~\cite{ribeiro:access:2025}
dataset and of the corresponding software repositories listed in
\secref{sec:intro:contributions}. The combination of these
contributions provides a unified \gls{slam} solution that ties
together distance map-based registration,
dynamics-aware local mapping, and offline global refinement,
addressing in an integrated manner to the two research questions related to
improving localization and mapping process in dynamic environments
formulated in \secref{sec:intro:rq+h}.

\section{Future Work and Final Remarks}\label{sec:conclusions:final-remarks}

This PhD~Thesis advanced the state of the art in 2D~\gls{slam} for
dynamic indoor environments along several complementary directions,
from feature extraction and 2D~point cloud registration to dynamics-aware
\gls{slam} and offline global refinement. The main results discussed in
\secref{sec:conclusions:contributions} validate the hypotheses formulated in
\secref{sec:intro:rq+h} and provide a basis for further developments.
Nevertheless, the research also exposes a set of open problems and natural
extensions that were identified during the development of this PhD~Thesis,
but whose execution was beyond its scope. This section consolidates the
directions of future work and provides an overview of the candidate's
involvement in the scientific community throughout the PhD~Thesis.

\subsection{Future work}\label{sec:conclusions:future-work}

The directions of future work identified in
\chapref{chap:features}, \chapref{chap:distance-maps}, and
\chapref{chap:slam} are consolidated as follows, ranging from the front-end
perception modules to the offline global refinement stages:
\begin{itemize}
\item \textbf{Covariance estimation for the corner-like features.}
The corner-like feature detector proposed in \chapref{chap:features}
produces the keypoint location as the closed-form intersection of two
locally fitted lines at multiple distance-based scales, but does not
yet provide an explicit uncertainty model for the feature observation.
Propagating the covariance of the line-fitting parameters through the
closed-form intersection step would yield a per-feature covariance
suitable for weighting the feature observations within a probabilistic
data association and tracking module. This extension is a prerequisite
for the integration of the corner-like features as observations within
a feature-based tracking module;
\item \textbf{Hybrid raw-scan, feature, and beacon-based \gls{slam}
framework.}
The current RicoSLAM framework focuses on the raw 2D~laser scanner data
as the only perceptual modality. A natural extension is the
development of a hybrid pose-landmark \gls{slam} framework that
simultaneously supports raw scan observations, geometric features
(\eg, line segments, the corner-like features proposed in
\chapref{chap:features}), and artificial beacons (\eg, reflective
cylinders). Such a framework would combine the robustness of dense
distance map-based registration with the long-range associability of
beacon-based and feature-based observations, while reusing the
$SE(2)$ graph back-end already adopted by RicoSLAM. The
covariance estimation extension for the corner-like features mentioned
above is a prerequisite for this development;
\item \textbf{Improvements on the precomputed point-to-plane
formulation.}
The precomputed point-to-plane variant
$\mathcal{D}_\mathrm{p2pl}(\mathcal{M}_2)$ achieves competitive
registration accuracy with respect to the \gls{icp} baselines, but
is more sensitive to range noise than the precomputed point-to-point
variant $\mathcal{D}_\mathrm{p2p}(\mathcal{M}_1)$, attributed to the
compounding of finite-difference errors when approximating both the
gradient and the Hessian of the distance field. Alternative strategies
to estimate these quantities, such as semi-analytical formulations that combine
the analytical gradient at $\mathcal{M}_1$ with a smoothed Hessian, or
truncated-distance reformulations that explicitly handle the
truncation boundaries, may reduce this sensitivity and improve the
robustness of the precomputed point-to-plane variant under noise.
\item \textbf{Improvements on the uncertainty estimation analysis for
2D~point cloud registration.}
The pose-error consistency metric
$\chi^2_{\boldsymbol{\varepsilon}}$ adopted in
\secref{sec:distance-maps:results:synthetic} is an
approximation rather than a true Mahalanobis statistic, because the
target-frame pose error
$\boldsymbol{\varepsilon}\!\left(\mathbf{x}\right)$ derived from
$\mathbf{x}\cdot\mathbf{x}^{*,-1}$ and the body-frame information
matrix $\boldsymbol{\Omega}$ returned by the Gauss-Newton solver
under the right-sided perturbation update live on different tangent
spaces of $SE(2)$. Both representations only coincide, up to the
adjoint twist $\mathrm{Ad}_{\mathbf{x}^*}^T\,\boldsymbol{\Omega}\,
\mathrm{Ad}_{\mathbf{x}^*}$, in the small angular residual regime.
Nevertheless, the target-frame parametrization was preferred in this
PhD~Thesis because of its diagnostic value in degenerate scenes,
projecting otherwise unobservable residuals onto the observable directions of
$\boldsymbol{\Omega}$ and inflates the metric whenever the alignment
fails along a degenerate axis.
\item \textbf{Improvements on the online dynamics removal mechanism.}
The online dynamics removal stage of RicoSLAM operates on a
per-keyframe basis and commits the updated local map directly to the
back-end graph once a split or merge criterion is triggered. An
intermediate buffering stage between the tracker and the back-end,
which could accumulate evidence over a sliding window of keyframes
before committing them, may further reduce the influence of transient
returns on the back-end graph and improve the robustness of the
framework in highly dynamic scenarios. In addition, the simplified
single-return ray model adopted by the online and offline dynamics
removal mechanisms could be extended to multi-return formulations,
addressing the residual semi-static dynamics observed in the Avantgarde
sequences and improving the cleaning effect on the final occupancy
grid representation.
\item \textbf{Integration and real-world validation of the distance
map-based \gls{gba}.}
The distance map-based \gls{gba} factors proposed in
\secref{sec:slam:gba} were validated on a synthetic benchmark with
ground-truth poses, isolating their contribution from the online
operation of RicoSLAM. The full integration of the proposed \gls{gba}
stage with the online RicoSLAM pipeline and its end-to-end validation
in real-world environments stands out as the most immediate direction
of future work. This integration would close the loop between the
online graph \gls{slam} operation and the offline global refinement
stage, enabling the explicit measurement of the contribution of the
\gls{gba} stage to the trajectory accuracy and to the quality of the
2D~occupancy grid maps produced by RicoSLAM in real-world industrial
deployments.
\end{itemize}

\subsection{Involvement in the scientific community}\label{sec:conclusions:community}
TBC

\subsubsection{Collaborations}
Giorgio Sapienza, Rome (PhD Internship, 6 months, February - July 2024).

5dpo Robotics Team:
\begin{description}[leftmargin=1.25cm,style=nextline,font=\normalfont]
\item[\cite{sousa:icarsc:2021}] \fullcite{sousa:icarsc:2021}
\item[\cite{sousa:icarsc:2024}] \fullcite{sousa:icarsc:2024}
\item[\cite{lopes:icarsc:2025}] \fullcite{lopes:icarsc:2025}
\end{description}

\subsubsection{Scientific projects}
APPLICATION:
SLAM4all (SLAM em tempo real para multi-domínio com múltiplos robôs sob transmissão de dados limitada).

PROJECTS:
VINCI7D (Gerber parser);
GreenAuto;
PRODUTECH R3.

\begin{itemize}
% \item GreenAuto:
%   \begin{description}[leftmargin=1.25cm,style=nextline,font=\normalfont]
%   \end{description}
\item PRODUTECH~R3:
  \begin{description}[leftmargin=1.25cm,style=nextline,font=\normalfont]
  \item[\cite{levin:robot:2024}] \fullcite{levin:robot:2024}
  \item[\cite{caldana:robot:2024}] \fullcite{caldana:robot:2024}
  \item[\cite{pacheco:icarsc:2025}] \fullcite{pacheco:icarsc:2025}
  \end{description}
\item VINCI~7D:
  \begin{description}[leftmargin=1.25cm,style=nextline,font=\normalfont]
  \item[\cite{sousa:access:2022}] \fullcite{sousa:access:2022}
  \end{description}
\end{itemize}



\subsubsection{Supervision}
B.Sc. supervision (not official) at FEUP (February - June 2023; 2nd semester 2022/23):
Rodrigo de Vasconcelos e Miguel (Compensação do Movimento do Robô nos Dados de Sensores Exteroceptivos);
José Miguel Guedes (Controlo de Trajetórias de Robôs Móveis com o Hangfa Discovery Q2);
André Luís Pires (Robótica e Sistemas Autónomos em Localização e Mapeamento de Robôs Móveis com o 3-omni 5DPO MSL);
Ana Luísa Mota (Robótica e Sistemas Autónomos em Localização e Mapeamento de Robôs Móveis com o Hangfa Discovery Q2).

B.Sc. supervision (not official) at ISEP (March - July 2023; 2nd semester 2022/23):
Paulo Ricardo Barbosa (Teste de algoritmos de mapeamento open source recorrendo ao sensor RS-Helios-5515).

BII at INESC TEC:
Daniele da Mota Caldana (Aplicação para deteção de local para manipulação de paletes e paletes danificadas);
Duarte Filipe Monteiro (Caixas metalicas GreenAuto);
David José Castro (Caixas metalicas GreenAuto);
João Miguel Oliveira (tele-controlo AMRs);
Thiago Baldassarri Levin (comunicações 5G).
