\chapter{Point Cloud Registration with Distance Maps}\label{chap:distance-maps}

This chapter presents a novel approach for 2D~point cloud registration
using distance maps, developed within the scope of this PhD~Thesis
in collaboration with Prof.~Dr.~Giorgio~Grisetti from
Sapienza Università di Roma.
Distance maps encode the Euclidean distance to the nearest reference point
at every cell of a precomputed discrete grid, decoupling data association
from the iterative optimization loop.
By replacing the explicit nearest-neighbour search performed at every
iteration of classical
\glslink{icp}{\glsentryfull{icp}}~\cite{besl-mckay:tpami:1992,chen-medioni:icra:1991}
with a constant-time grid lookup on distance map-based data structures,
the proposed formulation provides both point-to-point and point-to-plane error
formulations compatible with Gauss-Newton optimization on the $SE(2)$ manifold.
The proposed approach does not require explicit data association at each iteration.

This chapter is organized as follows. \secref{sec:distance-maps:background}
covers point cloud registration methods from the literature and
the non-linear least-squares framework used throughout this work.
\secref{sec:distance-maps:alignment} formulates the proposed approach,
introducing the distance map representation,
the registration problem on the $SE(2)$ manifold, and both
point-to-point and point-to-plane error formulations with their Jacobians.
\secref{sec:distance-maps:results} presents experimental results on
synthetic and real-world data, evaluating registration accuracy and
computational performance.
Lastly, \secref{sec:distance-maps:conclusions} summarizes the main
conclusions and directions for future work.

\section{Background}\label{sec:distance-maps:background}

Point cloud registration is a fundamental problem in mobile robotics,
consisting of estimating the rigid-body transformation that best aligns
two point clouds acquired by a range sensor. This problem arises in a wide range
of applications, including \gls{slam}, odometry estimation, and map-based
localization. Thus, this section introduces the background underpinning
the proposed distance map-based registration approach.
\secref{sec:distance-maps:background:registration} reviews 2D~and
3D~point cloud registration methods from the literature,
ranging from classical \gls{icp} variants to probabilistic and
precomputed spatial gradient formulations, motivating the use of
distance maps as a unified and correspondence-free registration framework.
\secref{sec:distance-maps:background:least-squares} presents the
non-linear least-squares framework used throughout this chapter,
covering the Gauss-Newton algorithm, uncertainty estimation,
robust estimators, and the generalization to non-Euclidean state spaces.

\subsection{Point cloud registration}\label{sec:distance-maps:background:registration}
One of the most popular approaches for point cloud registration is
the \glslink{icp}{\glsentryfull{icp}} algorithm.
Besl and McKay~\cite{besl-mckay:tpami:1992} address registration of 3D~shapes
represented as point clouds, introducing the standard \gls{icp} algorithm.
The following two steps can summarize this algorithm:
(i) compute correspondences between the source ($\mathbf{p}_i$) and target ($\mathbf{z}_i$)
cloud points (closest nearest neighbor based on Euclidean distance); and
(ii) estimate the rigid transformation of the source \gls{wrt} target cloud
($\mathbf{x}$) that minimizes a point-to-point distance measure between
corresponding points (see \eqref{eq:distance-maps:background:point-to-point}).
Chen and Medioni~\cite{chen-medioni:icra:1991} exploit the tendency of
most range data to be locally planar and introduce a point-to-plane variant
of \gls{icp}. The proposed method exploits surface normal information
on the target cloud ($\mathbf{n}_{z_i}$).
Instead of minimizing the difference between the projected source and
the target points, the point-to-plane error metric minimizes
the squared dot product of the surface normal with
the difference between the expected projected source and
the reference target points
(see \eqref{eq:distance-maps:background:point-to-plane}).
Indeed, Rusinkiewicz and Levoy~\cite{rusinkiewicz-levoy:im:2001} compared
\gls{icp} variants with respect to point selection,
data association, weighting of associated pairs, outlier rejection, and
error metrics. The point-to-plane~\cite{chen-medioni:icra:1991}
metric demonstrated significantly better performance than
the point-to-point~\cite{besl-mckay:tpami:1992} formulation in the experiments.
\begin{equation}\label{eq:distance-maps:background:point-to-point}
\mathbf{x} = \argmin_\mathbf{x} \sum_{i=1}^M \| \mathbf{x} \mathbf{p}_i - \mathbf{z}_i \|^2
\end{equation}
\begin{equation}\label{eq:distance-maps:background:point-to-plane}
\mathbf{x} = \argmin_\mathbf{x} \sum_{i=1}^M \| \mathbf{n}_{z_i}^T \cdot
\left( \mathbf{x} \mathbf{p}_i - \mathbf{z}_i \right) \|^2
\end{equation}

Unlike traditional \gls{icp} methods, which rely on explicit point
correspondences, the \glslink{ndt}{\glsentryfull{ndt}}
proposed by Biber and Strasser~\cite{biber-strasser:iros:2003} performs
2D~point cloud registration by representing the environment
with local normal distributions. The 2D~space is discretized into cells.
Each cell is associated with a Gaussian distribution defined by
the mean and covariance of the cell points,
which provide information about the surface those points represent.
Then, Biber and Strasser~\cite{biber-strasser:iros:2003} formulate
2D~point cloud registration as the minimization of an error function
derived from these distributions, optimized using the Gauss-Newton algorithm.
Experimental results showed that \gls{ndt}~\cite{biber-strasser:iros:2003} was
able to map large areas without requiring odometry.
Magnusson~\etal~\cite{magnusson:jfr:2007} later extended \gls{ndt} to
3D~point cloud registration by subdividing space into voxels,
each modelled as a normal distribution estimated from
the points contained within it. The alignment is obtained by
evaluating the likelihood of points from one scan under
the distributions of the other scan.
Compared with point-to-point \gls{icp}~\cite{besl-mckay:tpami:1992} using
kd-tree nearest-neighbor search~\cite{bentley:cacm:1975},
3D-\gls{ndt}~\cite{magnusson:jfr:2007} achieves
lower computation cost by avoiding explicit correspondence search,
while also offering a more compact representation of 3D~point cloud
compared to raw data.

While leveraging
\gls{ndt}~\cite{biber-strasser:iros:2003,magnusson:jfr:2007}-like
representations of the scene, \gls{gicp}~\cite{segal:rss:2009} and
\gls{nicp}~\cite{serafin-grisetti:iros:2015} still retain the \gls{icp}
structure for cloud registration.
\gls{gicp}~\cite{segal:rss:2009} retains the standard \gls{icp}
structure for correspondence estimation, typically using Euclidean
closest-neighbor search accelerated with a kd-tree~\cite{bentley:cacm:1975}.
The main modification lies in the minimization objective function,
introducing a plane-to-plane error metric. \gls{gicp} models
local surface information around each point, estimated from
neighboring points via \gls{pca}. By accounting for the covariance of
the matched points, \gls{gicp} weights residuals based on
local geometric uncertainty. Experiments showed that
\gls{gicp}~\cite{segal:rss:2009} was more robust to incorrect correspondences
and more accurate than standard point-to-point and point-to-plane
\gls{icp}~\cite{besl-mckay:tpami:1992,chen-medioni:icra:1991}.
\gls{nicp}~\cite{serafin-grisetti:iros:2015} extends
\gls{gicp}~\cite{segal:rss:2009} not only in the optimization stage but also
during correspondence search.
Rather than relying purely on Euclidean nearest-neighbor matching,
\gls{nicp}~\cite{serafin-grisetti:iros:2015} uses a line-of-sight-based
association step and rejects candidate pairs when normals are not well defined,
when point distance exceeds a threshold, when curvature mismatch is too large,
or when the normals are insufficiently aligned. In the optimization step,
\gls{nicp}~\cite{serafin-grisetti:iros:2015} augments each point with
its normal and minimizes a Mahalanobis distance in this 6D space.
In the experiments, \gls{nicp}~\cite{serafin-grisetti:iros:2015} was more
accurate than \gls{gicp}~\cite{segal:rss:2009} and
\gls{ndt}~\cite{magnusson:jfr:2007}
registration algorithms on the evaluated datasets.

Furthermore, Montesano~\etal~\cite{montesano:iros:2005} present
a probabilistic scan matching method for 2D~point cloud registration.
The method~\cite{montesano:iros:2005} builds a set of possible correspondences
using an individual compatibility test based on the Mahalanobis distance
between pairs of points, while accounting for correspondence uncertainty.
Then, Montesano~\etal~\cite{montesano:iros:2005} maximize
the probability of scan alignment over the points correspondence distribution.
The algorithm~\cite{montesano:iros:2005} showed faster convergence
rates than standard \gls{icp}~\cite{besl-mckay:tpami:1992}.
Instead of selecting a single set of correspondences as
Montesano~\etal~\cite{montesano:iros:2005}, Censi~\cite{censi:icra:2006}
considers all possible correspondences and generates alignment hypotheses
through the \gls{ght}, while avoiding explicit point-to-point correspondences.
Censi~\cite{censi:icra:2006} computes alignment within
a probabilistic framework that incorporates a motion model and estimates
for registration uncertainty, accounting for surface normals.

Similarly, the GMapping~\cite{grisetti:tro:2007} and
Cartographer~\cite{hess:icra:2016} \gls{slam} algorithms use \gls{mle}
probabilistic formulations to estimate the scan alignment \gls{wrt} the map.
The scan matcher in GMapping~\cite{grisetti:tro:2007} seeks
the most likely robot pose by matching the current observation against
the map built so far. The \gls{mle} formulation estimates
the robot pose given the map, the current scan, and an initial pose guess.
This formulation performs a gradient-descent search over
the observation likelihood function defined on the grid map, based on the
beam-endpoint sensor model~\cite{thrun:book:2005}.
Cartographer~\cite{hess:icra:2016} optimizes the scan pose relative to
the current local submap using a Ceres~\cite{ceres:sw:2023}-based scan matcher.
This matcher estimates the pose that maximizes the occupancy probabilities of
the scan points projected onto the submap. The problem is formulated as
a non-linear least-squares optimization, using a smooth version of
the occupancy probability values associated with the projected scan points.

Precomputed spatial gradients may be leveraged for the 2D~point cloud
registration without requiring explicit point-to-point correspondences.
HectorSLAM~\cite{kohlbrecher:ssr:2011} uses spatial gradients over the
2D~occupancy grid map ($\mathcal{M}$).
The scan matching finds the pose that maximizes
the occupancy probability of where the beam endpoints $\mathbf{z}_i$ land on occupied
cells (see \eqref{eq:distance-maps:background:hectorslam}).
The method uses bilinear interpolation over the occupancy map and approximates
its spatial derivatives, applying then a Gauss-Newton update.
PerfectMatch~\cite{lauer:robocup:2006,sobreira:jint:2019} focuses on
a soccer field markings environment for robot soccer.
This method formulates the 2D~registration problem based on distance maps, using
the \gls{rprop}~\cite{riedmiller-braun:icnn:1993} gradient-descent algorithm.
Each point is evaluated by the distance map ($\mathcal{D}_\mathcal{M}$)
that gives the distance to the closest field marking through a
robust M-estimator ($\rho$)
(see \eqref{eq:distance-maps:background:perfectmatch}).
The algorithm considers that the distance map is over the entire soccer field
environment, only considering scan-to-map alignment.
Although distance maps are continuous and thus differentiable,
the authors approximate the final estimation covariance based on the second
derivative of the error assuming an environment with orthogonal walls,
equivalent to the line markings of soccer fields.
Sobreira~\etal~\cite{sobreira:jint:2019} benchmarked in a simulation environment
PerfectMatch~\cite{lauer:robocup:2006} against
\gls{icp}~\cite{besl-mckay:tpami:1992} and
\gls{ndt}~\cite{biber-strasser:iros:2003}.
PerfectMatch~\cite{lauer:robocup:2006} was faster than both
\gls{icp}~\cite{besl-mckay:tpami:1992} and
\gls{ndt}~\cite{biber-strasser:iros:2003}, given that
PerfectMatch~\cite{lauer:robocup:2006} uses lookup tables
for the error function.
Additionally, the PerfectMatch~\cite{lauer:robocup:2006}
algorithm demonstrated a higher tolerance to orientation initialization errors
than \gls{icp}~\cite{besl-mckay:tpami:1992} and
\gls{ndt}~\cite{biber-strasser:iros:2003}.

\begin{equation}\label{eq:distance-maps:background:hectorslam}
\mathbf{x} = \argmin_\mathbf{x} \sum_{i=1}^M
\left[ 1 - \mathcal{M}\!\left(\mathbf{x} \mathbf{z}_i\right) \right]^2
\end{equation}
\begin{equation}\label{eq:distance-maps:background:perfectmatch}
\mathbf{x} = \argmin_\mathbf{x} \sum_{i=1}^M
\rho\!\left( \mathcal{D}_\mathcal{M}\!\left( \mathbf{x} \mathbf{z}_i \right) \right)
\end{equation}

Overall, most \gls{icp} methods rely on explicit point-to-point correspondences,
requiring two steps per optimization iteration: data association and
pairwise error computation between source and target points.
As for probabilistic-based approaches, these methods typically require solving a
\gls{map} problem and having a map estimate of the environment over time, which
is not compatible with scan-to-scan registration~\cite{segal:rss:2009}.
In contrast, distance maps
should be compatible with both scan-to-scan and scan-to-map alignment,
providing a nearest neighbor framework between source and target clouds.
While the PerfectMatch~\cite{lauer:robocup:2006} algorithm is specific
to scan-to-map registration, an adaptation of the original approach
should be compatible with scan-to-scan alignment.
Moreover, distance maps provide a precomputed lookup table
that directly allows evaluating the registration error of each source point
without requiring an explicit data association step.
This property allows faster computation during point cloud registration than
\gls{icp}~\cite{besl-mckay:tpami:1992} and
\gls{ndt}~\cite{biber-strasser:iros:2003}, as showed in
Sobreira~\etal~\cite{sobreira:jint:2019}.

% \subsection{Distance function representations}\label{sec:distance-maps:background:distance-functions}
%
% (surface reconstruction)
%
% signed distance function:
% - 10.1109/IROS.2015.7353633 (2D-SDF-SLAM,
%   only use SDF to improve map representation results,
%   occupancy probabilistic-based approach used for scan registration;
%   still, only works for scan-to-map alignment)
% - 2D: 10.1109/IROS40897.2019.8967683
% - 3D: 10.1109/LRA.2021.3052388
%
% TSDF: 10.1145/237170.237269 (original? incremental update to compute SDF)
% - 10.1109/CVPR.2015.7299077
% - Voxblox (10.1109/IROS.2017.8202315) / Voxgraph (10.1109/LRA.2019.2953859)
% - 10.1109/ICRA40945.2020.9196695

\subsection{Least-squares minimization}
\label{sec:distance-maps:background:least-squares}

State estimation in mobile robotics requires inferring an unknown state
from noisy sensor measurements~\cite{grisetti:itsm:2010,grisetti:robotics:2020}.
First, \secref{sec:distance-maps:background:least-squares:problem-formulation}
formulates the problem as a \gls{map} inference problem
and demonstrates how, under Gaussian noise assumptions, it reduces to
a non-linear least-squares problem. Next,
\secref{sec:distance-maps:background:least-squares:gn} shows how
the Gauss-Newton algorithm solves the non-linear least-squares problem
through a first-order approximation from the Taylor series.
\secref{sec:distance-maps:background:least-squares:uncertainty} presents
an analysis of the uncertainty of the recovered state estimate based on
the approximate Hessian matrix retrieved from the Gauss-Newton solver. Then,
\secref{sec:distance-maps:background:least-squares:robust-estimators} extends
the non-linear least-squares problem to use robust estimators to handle
outliers. Finally,
\secref{sec:distance-maps:background:least-squares:manifold} generalizes
the non-linear least-squares problems for non-Euclidean state spaces.

\subsubsection{Problem formulation}
\label{sec:distance-maps:background:least-squares:problem-formulation}

Least-squares minimization arises in robotics when estimating an unknown state
from noisy sensor measurements~\cite{grisetti:robotics:2020}.
Let $\mathbf{x}$ denote non-observable
state variables and $\mathbf{z}$ the measurements.
In the scope of this PhD~Thesis, the measurements in this chapter are the observations of the
environment acquired by the robot's on-board 2D~laser scanner.
The state is assumed to follow a prior distribution $p(\mathbf{x})$,
while the conditional distribution of the measurements given the state,
$p(\mathbf{z}|\mathbf{x})$, is assumed known,
commonly referred to as the \emph{observation model}. Then, the goal
is to estimate the most probable state given the measurements,
\ie, the posterior $p(\mathbf{x}|\mathbf{z})$.
This problem represents a \gls{map} estimation problem, as shown in
\eqref{eq:distance-maps:background:least-squares:map}~\cite{barfoot:book:2017}.
\begin{equation}\label{eq:distance-maps:background:least-squares:map}
\mathbf{x}^* = \argmax_\mathbf{x} p(\mathbf{x}|\mathbf{z})
\end{equation}

By applying Bayes' rule, shown in
\eqref{eq:distance-maps:background:least-squares:bayes}, the posterior
distribution $p(\mathbf{x}|\mathbf{z})$ becomes proportional to
the likelihood of the measurements given the state
$p(\mathbf{z}|\mathbf{x})$ multiplied by the prior $p(\mathbf{x})$.
If a uniform prior over the admissible values of $\mathbf{x}$ is assumed,
then $p(\mathbf{x})$ is constant and the
posterior $p(\mathbf{x}|\mathbf{z})$ depends only on the measurement
distribution conditioned by the state $p(\mathbf{z}|\mathbf{x})$.
\begin{equation}\label{eq:distance-maps:background:least-squares:bayes}
p(\mathbf{x}|\mathbf{z}) =
\frac{p(\mathbf{z}|\mathbf{x})p(\mathbf{x})}{p(\mathbf{z})}
\propto p(\mathbf{z}|\mathbf{x})p(\mathbf{x})
\propto p(\mathbf{z}|\mathbf{x})
\end{equation}

Next, considering $M$ measurements $\mathbf{z}^{[i]}$ that are
conditionally independent given $\mathbf{x}$, the likelihood
$p(\mathbf{x}|\mathbf{z})$ results in
\eqref{eq:distance-maps:background:least-squares:independence}.
Under the assumption that each measurement follows a Gaussian distribution
with mean $\mu_{z|x}^{[i]}=\mathbf{h}^{[i]}(\mathbf{x})$
(predicted measurement given the state $\mathbf{x}$) and covariance
$\mathbf{\Sigma}^{[i]} = {\mathbf{\Omega}^{[i]}}^{-1}$ (where
$\mathbf{\Omega}^{[i]}$ represents the information matrix of each measurement),
the individual likelihoods $p(\mathbf{z}^{[i]}|\mathbf{x})$ result in
\eqref{eq:distance-maps:background:least-squares:measurement}, based on the
probability density function of normal distributions.
\begin{equation}\label{eq:distance-maps:background:least-squares:independence}
p(\mathbf{x}|\mathbf{z}) \propto \prod_{i=1}^M p(\mathbf{z}^{[i]}|\mathbf{x})
\end{equation}
\begin{equation}\label{eq:distance-maps:background:least-squares:measurement}
\begin{aligned}
p(\mathbf{z}^{[i]}|\mathbf{x})
  & = \mathcal{N}(\mathbf{z}^{[i]} ; \mathbf{h}^{[i]}(\mathbf{x}),
                                    {\mathbf{\Omega}^{[i]}}^{-1})\\
  & \propto \exp\!\left(
      -\frac{1}{2}
      \left( \mathbf{h}^{[i]}(\mathbf{x}) - \mathbf{z}^{[i]} \right)^T
      \mathbf{\Omega}^{[i]}
      \left( \mathbf{h}^{[i]}(\mathbf{x}) - \mathbf{z}^{[i]} \right)
  \right)
\end{aligned}
\end{equation}

Furthermore, \eqref{eq:distance-maps:background:least-squares:error} defines
the residual error $\mathbf{e}^{[i]}(\mathbf{x})$ for each measurement.
The squared Mahalanobis distance ${\chi^2}^{[i]}(\mathbf{x})$ of the residual
error is formulated in \eqref{eq:distance-maps:background:least-squares:chi2}.
Thus, \eqref{eq:distance-maps:background:least-squares:measurement-detailed}
illustrates the proportionality of the individual likelihoods
$p(\mathbf{z}^{[i]}|\mathbf{x})$ over the squared Mahalanobis distance
${\chi^2}^{[i]}(\mathbf{x})$.
\begin{equation}\label{eq:distance-maps:background:least-squares:error}
\mathbf{e}^{[i]}(\mathbf{x})=\mathbf{h}^{[i]}(\mathbf{x}) - \mathbf{z}^{[i]}
\end{equation}
\begin{equation}\label{eq:distance-maps:background:least-squares:chi2}
{\chi^2}^{[i]}(\mathbf{x}) =
\mathbf{e}^{[i]}(\mathbf{x})^T \mathbf{\Omega}^{[i]} \mathbf{e}^{[i]}(\mathbf{x})
\end{equation}
\begin{equation}
\label{eq:distance-maps:background:least-squares:measurement-detailed}
\begin{aligned}
p(\mathbf{z}^{[i]}|\mathbf{x})
  & \propto \exp\!\left(
      -\frac{1}{2}
      \mathbf{e}^{[i]}(\mathbf{x})^T
      \mathbf{\Omega}^{[i]}
      \mathbf{e}^{[i]}(\mathbf{x})
  \right)\\
  & \propto \exp\!\left(-\frac{1}{2}{\chi^2}^{[i]}(\mathbf{x})\right)
\end{aligned}
\end{equation}

Under the assumption of having a uniform prior,
the original \gls{map} problem formulated in
\eqref{eq:distance-maps:background:least-squares:map} can be rewritten as
an \gls{mle} problem that maximizes the probability of the measurements given
the state $p(\mathbf{z}^{[i]}|\mathbf{x})$
(see \eqref{eq:distance-maps:background:least-squares:mle-log-likelihood}).
Through the negative log-likelihood
(see \defref{def:literature:slam:log-likelihood}), dropping constant factors,
and assuming the measurements under Gaussian noise,
the original \gls{map} problem for robotics state estimation reduces to a
\glslink{mle}{\glsentryfull{mle}} problem, which in turn reduces to minimizing
the sum of squared Mahalanobis residuals.
This problem is equivalent to a weighted non-linear least-squares problem,
as formalized in
\eqref{eq:distance-maps:background:least-squares:mle-log-likelihood}.
\begin{equation}\label{eq:distance-maps:background:least-squares:mle-log-likelihood}
\begin{aligned}
\mathbf{x}^*
  & = \argmax_\mathbf{x} \prod_{i=1}^M p(\mathbf{z}^{[i]}|\mathbf{x})\\
  & = \argmax_\mathbf{x} \prod_{i=1}^M \exp\!\left( -\frac{1}{2}\mathbf{e}^{[i]}(\mathbf{x})^T \mathbf{\Omega}^{[i]} \mathbf{e}^{[i]}(\mathbf{x}) \right)\\
  & = \argmin_\mathbf{x} \sum_{i=1}^M \mathbf{e}^{[i]}(\mathbf{x})^T \mathbf{\Omega}^{[i]} \mathbf{e}^{[i]}(\mathbf{x})\\
  & = \argmin_\mathbf{x} \sum_{i=1}^M {\chi^2}^{[i]}(\mathbf{x})
\end{aligned}
\end{equation}

\subsubsection{Gauss-Newton algorithm}
\label{sec:distance-maps:background:least-squares:gn}

Following the non-linear least-squares formulation commonly used in
robotics state estimation, one possible solution is an iterative minimization of
\eqref{eq:distance-maps:background:least-squares:iterative-function}~\cite{grisetti:robotics:2020}.
At each iteration $k$, the state is updated by a
perturbation $\mathbf{\Delta x}_k$, as shown in
\eqref{eq:distance-maps:background:least-squares:perturbation}.
\begin{equation}\label{eq:distance-maps:background:least-squares:iterative-function}
F\!\left(\mathbf{x}_k\right) =
\sum_{i=1}^M \mathbf{e}^{[i]}\!\left(\mathbf{x}_k\right)^T
               \mathbf{\Omega}^{[i]}
               \mathbf{e}^{[i]}\!\left(\mathbf{x}_k\right)
\end{equation}
\begin{equation}\label{eq:distance-maps:background:least-squares:perturbation}
\mathbf{x}_{k+1} = \mathbf{x}_k + \mathbf{\Delta x}_k
\end{equation}

In order to compute $\mathbf{\Delta x}_k$,
each residual is linearized around the current estimate $\mathbf{x}_k$
via a first-order Taylor approximation, as formulated in
\eqref{eq:distance-maps:background:least-squares:error-function-taylor}~\cite{barfoot:book:2017,nocedal-wright:book:2006},
where $\mathbf{J}^{[i]}\in\mathbb{R}^{m_i\times n}$ is the Jacobian of the
$i$-th residual evaluated at $\mathbf{x}_k$, with $m_i$ the dimension of the
residual $i$ and $n$ the dimension of the state.
\begin{equation}
\label{eq:distance-maps:background:least-squares:error-function-taylor}
\mathbf{e}^{[i]}\!\left(\mathbf{x}_k + \mathbf{\Delta x}_k\right)
\simeq
\underbrace{\mathbf{e}^{[i]}\!\left(\mathbf{x}_k\right)}_{\mathbf{e}^{[i]}} +
\underbrace{\left.\frac{\partial \mathbf{e}^{[i]}\!\left(\mathbf{x}\right)}{\partial \mathbf{x}}
\right|_{\mathbf{x}=\mathbf{x}_k}}_{\mathbf{J}^{[i]}}
\mathbf{\Delta x}_k
\end{equation}

Next, \eqref{eq:distance-maps:background:least-squares:products-expansion}
substitutes the first-order Taylor approximation into the objective function and
expands the products in
\eqref{eq:distance-maps:background:least-squares:iterative-function},
where the scalar symmetry
$\mathbf{\Delta x}^T {\mathbf{J}^{[i]}}^T \mathbf{\Omega}^{[i]}
\mathbf{e}^{[i]} = {\mathbf{e}^{[i]}}^T \mathbf{\Omega}^{[i]}
\mathbf{J}^{[i]} \mathbf{\Delta x}$ has been used to combine the two
cross-terms. Also, note that $\mathbf{b}^{[i]} = {\mathbf{J}^{[i]}}^T
\mathbf{\Omega}^{[i]} \mathbf{e}^{[i]} \in \mathbb{R}^{n}$ is defined as
a column vector. The notation ${\mathbf{b}^{[i]}}^T \mathbf{\Delta x}$ in
the expansion below reflects the transpose of that column vector.
\begin{equation}
\label{eq:distance-maps:background:least-squares:products-expansion}
\begin{aligned}
\mathbf{e}^{[i]}\!\left(\mathbf{x}_k + \mathbf{\Delta x}_k\right)^T
\mathbf{\Omega}^{[i]}
\mathbf{e}^{[i]}\!\left(\mathbf{x}_k + \mathbf{\Delta x}_k\right)
& \simeq \left(\mathbf{e}^{[i]} + \mathbf{J}^{[i]} \mathbf{\Delta x}\right)^T
         \mathbf{\Omega}^{[i]}
         \left(\mathbf{e}^{[i]} + \mathbf{J}^{[i]} \mathbf{\Delta x}\right)\\
& = \mathbf{\Delta x}^T \underbrace{{\mathbf{J}^{[i]}}^T \mathbf{\Omega}^{[i]} \mathbf{J}^{[i]}}_{\mathbf{H}^{[i]}} \mathbf{\Delta x} +
    2 \underbrace{{\mathbf{e}^{[i]}}^T \mathbf{\Omega}^{[i]} \mathbf{J}^{[i]}}_{{\mathbf{b}^{[i]}}^T} \mathbf{\Delta x} +
    \underbrace{{\mathbf{e}^{[i]}}^T \mathbf{\Omega}^{[i]} \mathbf{e}^{[i]}}_{c^{[i]}}
\end{aligned}
\end{equation}

Summing over all measurements and defining
$\mathbf{H} = \sum_{i=1}^M \mathbf{H}^{[i]} \in \mathbb{R}^{n \times n}$,
$\mathbf{b} = \sum_{i=1}^M \mathbf{b}^{[i]} \in \mathbb{R}^{n}$, and
$c = \sum_{i=1}^M c^{[i]} \in \mathbb{R}$,
\eqref{eq:distance-maps:background:least-squares:iterative-function-linearized}
formulates the linearized cost derived in
\eqref{eq:distance-maps:background:least-squares:products-expansion}
to a quadratic form in $\mathbf{\Delta x}$.
The information matrices of the measurements $\mathbf{\Omega}^{[i]}$ are
symmetric positive semi-definite. Consequently, the matrix $\mathbf{H}$
inherits the same property by construction, where
$\mathbf{H}^T=\mathbf{H}$~\cite{grisetti:robotics:2020}.
\begin{equation}
\label{eq:distance-maps:background:least-squares:iterative-function-linearized}
F\!\left(\mathbf{x}_k + \mathbf{\Delta x}_k\right)
\simeq
\mathbf{\Delta x}_k^T \mathbf{H} \mathbf{\Delta x}_k +
2 \mathbf{b}^T \mathbf{\Delta x}_k +
c
\end{equation}

The optimal perturbation minimizes the quadratic approximation presented in
\eqref{eq:distance-maps:background:least-squares:iterative-function-linearized}.
Setting the gradient \gls{wrt} $\mathbf{\Delta x}$ to zero, as shown in
\eqref{eq:distance-maps:background:least-squares:perturbation-minimization-solution},
yields the normal equations presented in
\eqref{eq:distance-maps:background:least-squares:b-hx}, whose solution
$\mathbf{\Delta x}_k$ is then used to update the state via
\eqref{eq:distance-maps:background:least-squares:perturbation}, constituting
one iteration of the \emph{Gauss-Newton} method.
The method linearizes the residual error functions around the current estimate
and minimizes the resulting quadratic approximation of the weighted non-linear
least-squares cost~\cite{grisetti:robotics:2020}.
\algref{alg:distance-maps:background:least-squares:gauss-newton} details the
steps for a single iteration of the minimization procedure.
\begin{equation}
\label{eq:distance-maps:background:least-squares:perturbation-minimization-solution}
\mathbf{0} =
\frac{\partial}{\partial \mathbf{\Delta x}}
\left[
  \mathbf{\Delta x}_k^T \mathbf{H} \mathbf{\Delta x}_k
  + 2 \mathbf{b}^T \mathbf{\Delta x}_k + c
\right] = 2\mathbf{H} \mathbf{\Delta x}_k + 2\mathbf{b}
\end{equation}
\begin{equation}\label{eq:distance-maps:background:least-squares:b-hx}
\mathbf{H} \mathbf{\Delta x}_k = -\mathbf{b}
\end{equation}

\input{algorithms/distance-maps/gauss-newton.tex}

\subsubsection{Least-squares and uncertainty}
\label{sec:distance-maps:background:least-squares:uncertainty}

At convergence, the approximate Hessian matrix $\mathbf{H}$ of the Gauss-Newton
algorithm provides an estimate of the uncertainty for the recovered state.
Consider a small perturbation $\mathbf{\Delta x}$ away from
the optimum $\mathbf{x}^*$.
Under the first-order linearization of each measurement function
$\mathbf{h}^{[i]}$, the estimated measurement
$\mathbf{h}^{[i]}\!\left(\mathbf{x}^*+\mathbf{\Delta x}\right)$ around the optimum
can be approximated by
\eqref{eq:distance-maps:background:least-squares:measurement-approximation},
where $\mathbf{J}^{[i]}\in\mathbb{R}^{m_i \times n}$ is the
Jacobian of the $i$-th measurement function evaluated at the optimum,
$m_i$ is the dimension of the $i$-th measurement, and
$n$ is the dimension of the state.
This linearization is valid as long as the measurement function is
sufficiently smooth and $\|\mathbf{\Delta x}\|$ is small.
\begin{equation}
\label{eq:distance-maps:background:least-squares:measurement-approximation}
\mathbf{h}^{[i]}\!\left(\mathbf{x}^* + \mathbf{\Delta x}\right)
\simeq
\mathbf{h}^{[i]}\!\left(\mathbf{x}^*\right)
+
\underbrace{
  \left.
    \frac{\partial \mathbf{h}^{[i]}(\mathbf{x})}{\partial \mathbf{x}}
  \right|_{\mathbf{x}=\mathbf{x}^*}
}_{\mathbf{J}^{[i]}}
\mathbf{\Delta x}
\end{equation}

Next,
\eqref{eq:distance-maps:background:least-squares:perturbated-state-conditional}
represents the conditional probability of the $i$-th measurement given the
perturbed state, where ${\mathbf{z}^{[i]}}^* = \mathbf{h}^{[i]}(\mathbf{x}^*)$
is the predicted measurement at the optimum.
\begin{equation}
\label{eq:distance-maps:background:least-squares:perturbated-state-conditional}
p(\mathbf{z}^{[i]}|\mathbf{x}^* + \mathbf{\Delta x})
= \mathcal{N}\!\left(
    \mathbf{z}^{[i]};\;
    \mu_{z|x}^{[i]}={\mathbf{z}^{[i]}}^* + \mathbf{J}^{[i]}\mathbf{\Delta x},
    {\mathbf{\Omega}^{[i]}}^{-1}
  \right)
\end{equation}

Since measurements are conditionally independent given the state,
the joint conditional likelihood
$p(\mathbf{z}|\mathbf{x}^*+\mathbf{\Delta x})$ over all $M$ measurements
factorizes into a product of Gaussian distributions of
each individual measurement. As a result, the joint conditional likelihood
$p(\mathbf{z}|\mathbf{x}^* + \mathbf{\Delta x})$ formulated in
\eqref{eq:distance-maps:background:least-squares:joint-perturbated-state-conditional}
is also a Gaussian over the stacked mean and the block-diagonal information
matrix, where
$\mathbf{J} = \left[{\mathbf{J}^{[1]}}^T, \ldots, {\mathbf{J}^{[M]}}^T\right]^T\in \mathbb{R}^{m \times n}$
is the stacked Jacobian, $m = \sum_{i=1}^{M} m_i$, and
$\mathbf{\Omega}_z = \mathrm{diag}\!\left(\mathbf{\Omega}^{[1]}, \ldots, \mathbf{\Omega}^{[M]}\right)$
is the block-diagonal information matrix reflecting the conditional independence
of the measurements.
\begin{equation}
\label{eq:distance-maps:background:least-squares:joint-perturbated-state-conditional}
p(\mathbf{z}|\mathbf{x}^* + \mathbf{\Delta x})
= \mathcal{N}\!\left(\mathbf{z};
    \mu_{z|x} = \mathbf{z}^* + \mathbf{J} \mathbf{\Delta x},
    \mathbf{\Omega}_z^{-1}
  \right)
\end{equation}

Assuming an uninformative (uniform) prior on $\mathbf{\Delta x}$ around the
optimum $\mathbf{x}^*$, this corresponds to a zero information matrix with
a zero mean for the $\mathbf{\Delta x}$, as shown in
\eqref{eq:distance-maps:background:least-squares:joint-perturbated-state-distribution}.
The distribution over the state perturbation $\mathbf{\Delta x}$ is not
a separate variable, given that $\mathbf{\Delta x} = \mathbf{x} - \mathbf{x}^*$.
Applying the Gaussian chain rule~\cite{barfoot:book:2017},
\eqref{eq:distance-maps:background:least-squares:joint-information-matrix}
defines the information matrix $\mathbf{\Omega}_{x,z}$
for the joint distribution over
$(\mathbf{\Delta x}, \mathbf{z})$.
The prior contributes no information to the top-left block, so
$\mathbf{\Omega}_{xx} = \mathbf{J}^T \mathbf{\Omega}_z \mathbf{J}$.
\begin{equation}
\label{eq:distance-maps:background:least-squares:joint-perturbated-state-distribution}
p(\mathbf{\Delta x}) = \mathcal{N}\!\left(\mathbf{\Delta x};\mathbf{0},\mathbf{\Omega}_{x}^{-1}\right),
\; \mathbf{\Omega}_{x} = \mathbf{0}
\end{equation}
\begin{equation}
\label{eq:distance-maps:background:least-squares:joint-information-matrix}
\mathbf{\Omega}_{x,z} =
\begin{pmatrix}
  \mathbf{\Omega}_{x} + \mathbf{J}^T \mathbf{\Omega}_{z} \mathbf{J} &
  \mathbf{J}^{T}\mathbf{\Omega}_{z}\\
  \mathbf{\Omega}_{z}\mathbf{J} &
  \mathbf{\Omega}_{z}
\end{pmatrix} =
\begin{pmatrix}
  \mathbf{J}^T \mathbf{\Omega}_{z} \mathbf{J} &
  \mathbf{J}^{T}\mathbf{\Omega}_{z}\\
  \mathbf{\Omega}_{z}\mathbf{J} &
  \mathbf{\Omega}_{z}
\end{pmatrix} =
\begin{pmatrix}
  \mathbf{\Omega}_{xx} & \mathbf{\Omega}_{xz} \\
  \mathbf{\Omega}_{zx} & \mathbf{\Omega}_{zz}
\end{pmatrix}
\end{equation}

When conditioning the state posterior $\mathbf{x}$ on the measurements $\mathbf{z}$ from the joint
probability and under the Gaussian assumption,
Sch\"{o}n and Lindsten~\cite{schon-lindsten:tech-report:2011}
demonstrate that the mean and the covariance matrix for the state posterior
given the measurements can be formulated by
\eqref{eq:distance-maps:background:least-squares:state-posterior-mean}
and
\eqref{eq:distance-maps:background:least-squares:state-posterior-covariance},
respectively. Since at the optimum state the residual
$\mathbf{z} - \mathbf{z}^* \simeq \mathbf{0}$, the correction term vanishes
in the mean of the state posterior given the measurements.
\begin{equation}
\label{eq:distance-maps:background:least-squares:state-posterior-mean}
\begin{aligned}
\mu_{x|z} & = \mu_x -
              \mathbf{\Omega}_{xx}^{-1} \mathbf{\Omega}_{xz}
              \left(\mathbf{z} - \mu_z\right)\\
& = \mathbf{x}^* -
    \mathbf{\Omega}_{xx}^{-1} \mathbf{\Omega}_{xz}
    \left(\mathbf{z} - \mathbf{z}^*\right)\\
& \simeq \mathbf{x}^*
\end{aligned}
\end{equation}
\begin{equation}
\label{eq:distance-maps:background:least-squares:state-posterior-covariance}
\begin{aligned}
\mathbf{\Sigma}_{x|z} & =
\mathbf{\Sigma}_{xx} - \mathbf{\Sigma}_{xz} \mathbf{\Sigma}_{zz}^{-1} \mathbf{\Sigma}_{zx}\\
& = \mathbf{\Omega}_{xx}^{-1}
\end{aligned}
\end{equation}

Therefore, the covariance of the optimized state $\mathbf{x}^*$ is estimated by
\eqref{eq:distance-maps:background:least-squares:state-covariance}, where
$\mathbf{\Omega}_{xx} = \mathbf{J}^T \mathbf{\Omega}_z \mathbf{J} = \mathbf{H}$
is the approximate Hessian recovered from the Gauss-Newton solver.
The approximation sign reflects the fact that $\mathbf{H}$ is itself a
first-order approximation to the true Hessian of the cost function,
obtained by discarding the second-order residual terms in the linearization.
The diagonal entries of $\mathbf{\Sigma}_{x|z}$ represent the marginal
variances of each state component, and the off-diagonal entries encode
the correlations induced by the measurement model.
This result requires $\mathbf{H}$ to have full column rank, \ie, the stacked
Jacobian $\mathbf{J}$ must be full column rank.
When this condition does not hold (\eg, when some state components are
unobservable given the available measurements), $\mathbf{H}$ is singular and
the covariance is undefined for those directions~\cite{barfoot:book:2017}.
\begin{equation}
\label{eq:distance-maps:background:least-squares:state-covariance}
\mathbf{\Sigma}_{x|z}
= \mathbf{\Omega}_{xx}^{-1}
= \left(\mathbf{J}^T \mathbf{\Omega}_z \mathbf{J}\right)^{-1}
\approx \mathbf{H}^{-1}
\end{equation}

\subsubsection{Robust estimators}
\label{sec:distance-maps:background:least-squares:robust-estimators}

The weighted non-linear least-squares algorithm formulated in
\eqref{eq:distance-maps:background:least-squares:mle-log-likelihood}
assumes that all measurements follow a Gaussian noise model.
Under this assumption, the cost
$F(\mathbf{x}) = \sum_{i=1}^{M} {\chi^2}^{[i]}(\mathbf{x})$
grows quadratically with the residual error $\mathbf{e}^{[i]}(\mathbf{x})$,
meaning that a single outlying measurement with a large ${\chi^2}^{[i]}$ can
dominate the sum and severely corrupt the optimum.
M-estimators address this limitation by replacing the quadratic penalty with
a robust loss function $\rho(\cdot)$ that down-weights large residuals,
resulting in the robust cost $F_{\rho}(\mathbf{x})$ presented in
\eqref{eq:distance-maps:background:least-squares:robust-cost},
where $s^{[i]}$ is the scalar Mahalanobis distance of the $i$-th
measurement~\cite{huber:book:1981}.
\begin{equation}
\label{eq:distance-maps:background:least-squares:robust-cost}
F_{\rho}(\mathbf{x}) =
\sum_{i=1}^{M} \rho\!\left(s^{[i]}(\mathbf{x})\right),
\;
s^{[i]}(\mathbf{x}) = \sqrt{{\chi^2}^{[i]}(\mathbf{x})}
\end{equation}

The standard least-squares cost $F(\mathbf{x})$ is recovered as
a special case of $F_{\rho}(\mathbf{x})$ when $\rho(s) = s^2$.
Consider the gradient of the standard least-squares with respect to $\mathbf{x}$
presented in
\eqref{eq:distance-maps:background:least-squares:standard-gradient},
using the identity
$\frac{\partial}{\partial \mathbf{x}}\!\left(\mathbf{e}^{[i]}(\mathbf{x})^T \mathbf{\Omega}^{[i]} \mathbf{e}^{[i]}(\mathbf{x})\right) = 2 \cdot {\mathbf{J}^{[i]}}^T \mathbf{\Omega}^{[i]} \mathbf{e}^{[i]}$
because $\mathbf{\Omega}^{[i]}$ is symmetric.
\begin{equation}
\label{eq:distance-maps:background:least-squares:standard-gradient}
\begin{aligned}
\nabla_{\mathbf{x}} F
& = \sum_{i=1}^{M}
    \frac{\partial}{\partial \mathbf{x}}
    \!\left(\mathbf{e}^{[i]}(\mathbf{x})^T \mathbf{\Omega}^{[i]} \mathbf{e}^{[i]}(\mathbf{x})\right)
  = \sum_{i=1}^{M} 2 \cdot {\mathbf{J}^{[i]}}^T \mathbf{\Omega}^{[i]} \mathbf{e}^{[i]}\\
& = \sum_{i=1}^{M} \nabla_{\mathbf{x}} F^{[i]}
\end{aligned}
\end{equation}

When expanding the gradient of the robust cost
$F_{\rho}(\mathbf{x})$ presented in
\eqref{eq:distance-maps:background:least-squares:robust-gradient},
the derivation chain rule is applied using the gradient identity from
\eqref{eq:distance-maps:background:least-squares:standard-gradient},
and the influence of the robust estimator $\rho$ appears as a scalar weight $w$
of each measurement factor in $F_{\rho}(\mathbf{x})$ when compared to
the gradient of the standard cost $F(\mathbf{x})$.
This influence results in a re-weighted least-squares problem,
introducing the effective information matrix for each measurement factor
$\mathbf{\tilde{\Omega}}^{[i]}(\mathbf{x}_k)$ formulated in
\eqref{eq:distance-maps:background:least-squares:effective-info}.
\begin{equation}
\label{eq:distance-maps:background:least-squares:robust-gradient}
\begin{aligned}
\nabla_{\mathbf{x}} F_{\rho} =
\sum_{i=1}^{M}
\frac{\partial}{\partial \mathbf{x}}
\!\left(\rho\!\left(s^{[i]}(\mathbf{x})\right)\right)
& = \sum_{i=1}^{M}
    \rho'\!\left(s^{[i]}(\mathbf{x})\right)
    \cdot
    \frac{\partial \left(\sqrt{{\chi^2}^{[i]}(\mathbf{x})}\right)}{\partial \mathbf{x}}
  = \sum_{i=1}^{M}
    \frac{\rho'\!\left(s^{[i]}(\mathbf{x})\right)}{2 \cdot s^{[i]}(\mathbf{x})}
    \cdot
    \nabla_{\mathbf{x}} F^{[i]}\\
& = \sum_{i=1}^{M}
    w(s^{[i]}(\mathbf{x}))
    \cdot
    \nabla_{\mathbf{x}} F^{[i]}, \;
    w(s) = \frac{\rho'(s)}{2s}
\end{aligned}
\end{equation}
\begin{equation}
\label{eq:distance-maps:background:least-squares:effective-info}
\mathbf{\tilde{\Omega}}^{[i]}(\mathbf{x}_k)
= w^{[i]}(\mathbf{x}_k)\,\mathbf{\Omega}^{[i]},
\end{equation}

Subsequently,
\eqref{eq:distance-maps:background:least-squares:robust-hb}
defines the robust approximate Hessian matrix $\mathbf{\tilde{H}}$ and
gradient vector $\mathbf{\tilde{b}}$ for the Gauss-Newton
method to solve least-squares problems with robust estimators.
The only modification with respect to the standard formulation is the
substitution $\mathbf{\Omega}^{[i]} \leftarrow \mathbf{\tilde{\Omega}}^{[i]}$,
so any sparse linear solver used for the standard Gauss-Newton step can be
reused without modification.
\algref{alg:distance-maps:background:least-squares:gauss-newton:robust} details
the steps for a single iteration of the Gauss-Newton minimization procedure
with robust estimators.
\begin{equation}
\label{eq:distance-maps:background:least-squares:robust-hb}
\mathbf{\tilde{H}}
= \sum_{i=1}^{M}
  {\mathbf{J}^{[i]}}^T
  \mathbf{\tilde{\Omega}}^{[i]}
  \mathbf{J}^{[i]},
\;
\mathbf{\tilde{b}}
= \sum_{i=1}^{M}
  {\mathbf{J}^{[i]}}^T
  \mathbf{\tilde{\Omega}}^{[i]}\,
  \mathbf{e}^{[i]}(\mathbf{x}_k).
\end{equation}

\input{algorithms/distance-maps/gauss-newton_robust.tex}

At convergence, the robust approximate Hessian $\mathbf{\tilde{H}}$ provides
a covariance estimate analogous to
\eqref{eq:distance-maps:background:least-squares:state-covariance}, as presented
in \eqref{eq:distance-maps:background:least-squares:robust-covariance}.
Given that outlier factors receive $w^{[i]} \approx 0$, they contribute
negligibly to $\mathbf{\tilde{H}}$, not inflating the reported uncertainty of
the inlier-driven solution~\cite{huber:book:1981}.
\begin{equation}
\label{eq:distance-maps:background:least-squares:robust-covariance}
\mathbf{\tilde{\Sigma}}_{x|z}
\approx \mathbf{\tilde{H}}^{-1}.
\end{equation}

Lastly, Table~\ref{tab:distance-maps:background:least-squares:robust-losses}
summarizes popular choices for the robust loss function $\rho$, together with
their first-order derivative  $\psi(s) = \rho'(s)$ and
weight function $w(s) = \psi(s)/(2s)$~\cite{babin:icra:2019}.
The tuning constant $c$ controls the threshold between the inlier region,
where the loss behaves near-quadratically, and the outlier region,
where large residuals are down-weighted.

\input{tables/distance-maps/robust-estimators.tex}

\subsubsection{Non-Euclidean spaces}
\label{sec:distance-maps:background:least-squares:manifold}

State estimation in robotics frequently involves state spaces that are not
flat Euclidean spaces. For example, robot poses include orientations that
are elements of the special orthogonal group $SO(n)$ ($SO(2)$ and $SO(3)$
for 2D~and 3D~rotation matrices), or full rigid-body transformations that
are elements of the special Euclidean group $SE(n)$, combining rotations
and translations in $\mathbb{R}^{n}$ ($SE(2)$ and $SE(3)$ for 2D~and 3D~
poses).
These groups are smooth manifolds that do not support
standard Euclidean addition and subtraction of state variables.
Thus, the geometry of the state space must be accounted for explicitly during
optimization~\cite{barfoot:book:2017,grisetti:robotics:2020}.

A smooth manifold $\mathcal{M}$ is a space that locally resembles a Euclidean
space $\mathbb{R}^n$. At each point $\mathbf{x}\in\mathcal{M}$,
the \emph{tangent space}
$\mathcal{T}_{\mathbf{x}}(\mathcal{M})\cong\mathbb{R}^{n}$ provides
the appropriate linear structure for computing increments,
required for iterative least-squares minimization.
A small perturbation $\mathbf{\Delta x}\in\mathbb{R}^{n}$ expressed
in the tangent space is mapped back onto the manifold through the
\emph{exponential mapping}, as formulated in
\eqref{eq:distance-maps:background:least-squares:manifold-exp-plus-operator}.
The operator $\oplus$ replaces standard Euclidean
addition throughout the optimization, ensuring that every iteration remains on
$\mathcal{M}$, while preserving geometric constraints intrinsic to
the state representation (\eg, the orthogonality constraint on rotation
matrices, or the unit-norm constraint on quaternions).
The inverse operation maps an element $\mathbf{y}\in\mathcal{M}$ back to
the tangent space at $\mathbf{x}$ through the \emph{logarithmic map},
as formulated in
\eqref{eq:distance-maps:background:least-squares:manifold-log-minus-operator}~\cite{barfoot:book:2017}.
In the Euclidean case ($\mathcal{M} = \mathbb{R}^n$), the operators $\oplus$ and
$\ominus$ reduce to standard vector addition and subtraction operations.
\begin{equation}
\label{eq:distance-maps:background:least-squares:manifold-exp-plus-operator}
\mathbf{x} \oplus \mathbf{\Delta x}
\triangleq
\mathrm{exp}_{\mathbf{x}}(\mathbf{\Delta x})
\in \mathcal{M}
\end{equation}
\begin{equation}
\label{eq:distance-maps:background:least-squares:manifold-log-minus-operator}
\mathbf{y} \ominus \mathbf{x}
\triangleq
\mathrm{log}_{\mathbf{x}}(\mathbf{y})
\in \mathcal{T}_{\mathbf{x}}(\mathcal{M})
\end{equation}

Given the current state estimate $\mathbf{x}_k\in\mathcal{M}$ and
a tangent-space perturbation $\mathbf{\Delta x}\in\mathbb{R}^{n}$,
the first-order Taylor expansion of the residual is formulated by
\eqref{eq:distance-maps:background:least-squares:manifold-linearization}.
\eqref{eq:distance-maps:background:least-squares:manifold-error} represents the
\emph{logarithmic map} of the predicted measurement
$\mathbf{h}^{[i]}(\mathbf{x}_k)$ into the tangent space at the reference
measurement $\mathbf{z}^{[i]}$.
The \emph{manifold Jacobian} (see
\eqref{eq:distance-maps:background:least-squares:manifold-jacobian})
is the derivative of the residual \gls{wrt}
$\mathbf{\Delta x}\in\mathbb{R}^{n}$, evaluated at the origin of the tangent
space (equivalent to $\mathbf{\Delta x} = 0$)~\cite{grisetti:robotics:2020}.
The choice of left versus right-sided perturbation update in Lie groups affects
the form of the Jacobian. Hence, \emph{exponential} and \emph{logarithmic map}
operations should be consistent throughout the implementations.
\begin{equation}
\label{eq:distance-maps:background:least-squares:manifold-linearization}
\mathbf{e}^{[i]}\!\left(\mathbf{x}_k \oplus \mathbf{\Delta x}\right)
\approx
\mathbf{e}^{[i]}(\mathbf{x}_k)
+ \mathbf{J}^{[i]}(\mathbf{x}_k) \mathbf{\Delta x}
\end{equation}
\begin{equation}
\label{eq:distance-maps:background:least-squares:manifold-error}
\mathbf{e}^{[i]}(\mathbf{x}_k) =
\mathbf{h}^{[i]}(\mathbf{x}_k) \ominus \mathbf{z}^{[i]}
\in \mathcal{T}_{\mathbf{z}^{[i]}}(\mathcal{M})
\end{equation}
\begin{equation}
\label{eq:distance-maps:background:least-squares:manifold-jacobian}
\mathbf{J}^{[i]}(\mathbf{x}) =
\left.
  \frac{\partial \mathbf{e}^{[i]}\!\left(\mathbf{x} \oplus \mathbf{\Delta x}\right)}
        {\partial \mathbf{\Delta x}}
\right|_{\mathbf{\Delta x}=\mathbf{0}}
\in \mathbb{R}^{m_i \times n}
\end{equation}

As a result, the definition of the approximate Hessian and gradient vectors
based on the Jacobian computed \gls{wrt} the tangent-space perturbation rather
than a Euclidean increment (see
\eqref{eq:distance-maps:background:least-squares:manifold-jacobian})
and the error from the \emph{logarithmic map} of the measurement and prediction
estimation make the iterative least-squares on manifolds similar to
\algref{alg:distance-maps:background:least-squares:gauss-newton} and
\algref{alg:distance-maps:background:least-squares:gauss-newton:robust},
differing only in how $\mathbf{H}$ and $\mathbf{b}$ are computed.
Once $\mathbf{\Delta x}^*$ has been computed in the tangent space through
standard linear solvers as in the original Gauss-Newton method, the state is
retracted back onto the manifold via the \emph{exponential map}, as formulated
in \eqref{eq:distance-maps:background:least-squares:manifold-update}.
\begin{equation}
\label{eq:distance-maps:background:least-squares:manifold-update}
\mathbf{x}_{k+1} = \mathbf{x}_k \oplus \mathbf{\Delta x}^*
= \mathrm{exp}_{\mathbf{x}_k}\!\left(\mathrm{solve}\!\left(
  -\mathbf{H}_k^{-1} \mathbf{b}_k
\right)\right)
\end{equation}

\section{Registration Methodology}
\label{sec:distance-maps:alignment}

This section presents the proposed approach for 2D~point cloud registration
using distance maps, formulated as a non-linear least-squares problem and
solved via the Gauss-Newton algorithm on the $SE(2)$ manifold.
First, \secref{sec:distance-maps:alignment:dmaps} introduces the distance map
representation, its spatial derivatives, the associated representation
levels, and the data structures used for computation and querying.
Next, \secref{sec:distance-maps:alignment:problem} states the registration
problem, defining the pose parametrization and the manifold update. Then,
\secref{sec:distance-maps:alignment:pt2pt} derives the point-to-point error
formulation and its Jacobian matrix via right-sided perturbation on $SE(2)$.
Lastly, \secref{sec:distance-maps:alignment:pt2pl} derives the point-to-plane
error formulation and its Jacobian, together with the surface normal
estimation procedure used to extract normals from the moving point cloud.

\subsection{Distance maps}
\label{sec:distance-maps:alignment:dmaps}

Distance maps, also referred in the literature as distance fields or transforms,
are scalar fields that associate to every cell of a discrete grid
the Euclidean distance to the nearest element of a given reference set.
In the context of point cloud registration, the reference set is
the target point cloud and the distance map is precomputed only once before
the iterative optimization begins, in contrast to classical
\gls{icp} methods~\cite{besl-mckay:tpami:1992,chen-medioni:icra:1991}.
This precomputation encodes data association implicitly,
removing the explicit \glslink{nn}{\glsentryfull{nn}} search that traditional
\gls{icp} methods~\cite{besl-mckay:tpami:1992,chen-medioni:icra:1991} perform at
every iteration and replacing it with a constant-time grid lookup.

Therefore, \secref{sec:distance-maps:alignment:dmaps:overview} defines the
distance map over the reference point cloud and discusses how it removes the
explicit data association step required by classical
\gls{icp} methods~\cite{besl-mckay:tpami:1992,chen-medioni:icra:1991}.
\secref{sec:distance-maps:alignment:dmaps:derivation} derives the first- and
second-order spatial derivatives of the distance map, providing both
analytical expressions and finite-difference approximations used in
the implementation of the proposed approach.
\secref{sec:distance-maps:alignment:dmaps:representation} describes the three
representation levels of increasing information content considered in this work.
Lastly, \secref{sec:distance-maps:alignment:dmaps:computation} details the
computation pipeline and the data structures underlying each variant of
the distance map representations considered in the scope of this PhD~Thesis.

\subsubsection{Overview}
\label{sec:distance-maps:alignment:dmaps:overview}

Let $\mathcal{P} = \{\mathbf{p}_1, \mathbf{p}_2, \dots, \mathbf{p}_n\} \subset \mathbb{R}^2$
represent a reference 2D~point cloud,
comprising $n$ 2D~points acquired by a planar laser scanner.
These points are expressed in the sensor coordinate frame $S$,
where each point is $\mathbf{p}_i = (x_i, y_i)^T$.
The distance map $\mathcal{D}: \mathbb{R}^2\to\mathbb{R}_0^+$ is defined
as the function that maps any query point $\mathbf{y}\in\mathbb{R}^2$ to
its Euclidean distance to the nearest point in $\mathcal{P}$, as formulated in
\eqref{eq:distance-maps:alignment:dmaps:definition},
where $\mathbf{p}^*(\mathbf{y}) = \argmin_{\mathbf{p}_i \in \mathcal{P}}
\|\mathbf{y} - \mathbf{p}_i\|_2$ is the \gls{nn} in
the reference cloud $\mathcal{P}$.
Note that $\mathcal{D}(\mathbf{y}) \geq 0$ for all $\mathbf{y}$, making
this an \emph{unsigned distance function}, as opposed to
\glspl{sdf} used in volumetric reconstruction. An important property of
\emph{unsigned distance fields} is that they are continuous everywhere,
but not everywhere differentiable~\cite{jones:tvcg:2006}.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:definition}
\mathcal{D}(\mathbf{y}) =
\{ \mathbb{R}^2\to\mathbb{R}_0^+: \| \mathbf{y} - \mathbf{p}^*(\mathbf{y}) \|_2\}
\end{equation}

As discussed in \secref{sec:distance-maps:background:registration},
classical \gls{icp}-based registration
approaches~\cite{besl-mckay:tpami:1992,chen-medioni:icra:1991} implement a
two-step iterative process: data association and optimization steps.
These methods require reevaluating the correspondences at every iteration,
since the pose estimate evolves and changes the transformed positions of the
moving points, potentially altering the data association results.
In contrast, the distance map formulation removes this coupling entirely.
The distance map $\mathcal{D}$ is precomputed over a discrete grid
with a given $\delta$ (m/px) resolution covering the environment from
the reference cloud $\mathcal{P}$ prior to the registration process.
Subsequently, this structure is queried during optimization without requiring
any further reference to $\mathcal{P}$. The structure $\mathcal{D}$ encodes
information on the \gls{nn} point from $\mathcal{P}$ in each cell of the
distance map. Data association is thus resolved implicitly within
the error function evaluation itself, transforming a moving point
by the current pose estimate. At the resulting point location,
looking up into $\mathcal{D}$ yields information
(\eg, distance, its derivatives) on the nearest reference point
in $O(1)$,  with no explicit correspondence search.

\subsubsection{Spatial derivatives}
\label{sec:distance-maps:alignment:dmaps:derivation}

While the distance map $\mathcal{D}$ is continuous everywhere but not everywhere
differentiable~\cite{jones:tvcg:2006}, its spatial derivatives
exist almost everywhere and are central to the registration error formulations
derived in \secref{sec:distance-maps:alignment:pt2pt} and
\secref{sec:distance-maps:alignment:pt2pl}.
This subsubsection derives the first- and second-order spatial derivatives
of $\mathcal{D}$, providing both analytical expressions and their
finite-difference approximations used in our implementation.

\paragraph{First-order derivative (gradient vector)}
The gradient $\bm{\nabla}\mathcal{D}(\mathbf{y}) \in \mathbb{R}^2$ of the
distance map at a query point $\mathbf{y}$ can be obtained analytically
wherever $\mathcal{D}$ is smooth, \ie, at points admitting a unique
\gls{nn} in $\mathcal{P}$. At such a point,
$\mathbf{p}^*(\mathbf{y})$ is unique and the gradient is a unit vector pointing
from the nearest reference point toward $\mathbf{y}$, as given in
\eqref{eq:distance-maps:alignment:dmaps:1st-deriv:analytical-derivation},
which results from differentiating
\eqref{eq:distance-maps:alignment:dmaps:definition} \gls{wrt} $\mathbf{y}$.
Indeed, the unit-norm property $\|\bm{\nabla}\mathcal{D}\|_2 = 1$ holds
almost everywhere for \emph{unsigned distance fields},
with the exception of points equidistant from two or more reference points in
$\mathcal{P}$, where the gradient is not defined
(\eg, at the center of a circumference).
This property represents the Eikonal equation, a fundamental property
of Euclidean distance fields~\cite{jones:tvcg:2006}.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:1st-deriv:analytical-derivation}
\bm{\nabla}\mathcal{D}(\mathbf{y}) =
\frac{\mathbf{y} - \mathbf{p}^*(\mathbf{y})}
    {\|\mathbf{y} - \mathbf{p}^*(\mathbf{y})\|_2},
\; \|\bm{\nabla}\mathcal{D}(\mathbf{y})\|_2 = 1
\end{equation}

In the discrete grid setting, the gradient is estimated numerically using
central finite differences along the $y_1$ and $y_2$ directions of
$\mathbf{y} = (y_1, y_2)^T$ on the distance map $\mathcal{D}$, as
formulated in
\eqref{eq:distance-maps:alignment:dmaps:1st-deriv:numerical-derivation},
with cell resolution $\delta > 0$ and one-sided (forward or backward)
differences used at grid borders.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:1st-deriv:numerical-derivation}
\left.\frac{\partial \mathcal{D}}{\partial y_1}\right|_{\mathbf{y}}
\approx
\frac{\mathcal{D}(y_1 + \delta,\, y_2) -
      \mathcal{D}(y_1 - \delta,\, y_2)}{2\delta},
\;
\left.\frac{\partial \mathcal{D}}{\partial y_2}\right|_{\mathbf{y}}
\approx
\frac{\mathcal{D}(y_1,\, y_2 + \delta) -
      \mathcal{D}(y_1,\, y_2 - \delta)}{2\delta}
\end{equation}

\paragraph{Second-order derivative (Hessian matrix)}

The Hessian matrix $\mathbf{H}_{\mathcal{D}}(\mathbf{y})\in\mathbb{R}^{2\times2}$
is the symmetric matrix of second-order partial derivatives of $\mathcal{D}$, as
formulated in \eqref{eq:distance-maps:alignment:dmaps:2nd-deriv:H}, given that
$\frac{\partial^2 \mathcal{D}}{\partial y_1 \partial y_2} = \frac{\partial^2 \mathcal{D}}{\partial y_2 \partial y_1}$.
This Hessian matrix represents also the Jacobian matrix's transpose of
the analytical gradient $\bm{\nabla}\mathcal{D}(\mathbf{y})$, as formulated in
\eqref{eq:distance-maps:alignment:dmaps:2nd-deriv:H-Jgrad}.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:2nd-deriv:H}
\mathbf{H}_{\mathcal{D}}(\mathbf{y}) =
\begin{bmatrix}
\frac{\partial^2 \mathcal{D}}{\partial y_1^2} &
\frac{\partial^2 \mathcal{D}}{\partial y_1 \partial y_2} \\
\frac{\partial^2 \mathcal{D}}{\partial y_2 \partial y_1} &
\frac{\partial^2 \mathcal{D}}{\partial y_2^2}
\end{bmatrix}_{\mathbf{y}} =
\begin{bmatrix}
\frac{(y_2 - p_y^*(\mathbf{y}))^2}{\|\mathbf{y} - \mathbf{p}^*(\mathbf{y})\|_2^3} &
-\frac{(y_1 - p_x^*(\mathbf{y}))(y_2 - p_y^*(\mathbf{y}))}{\|\mathbf{y} - \mathbf{p}^*(\mathbf{y})\|_2^3}\\
-\frac{(y_1 - p_x^*(\mathbf{y}))(y_2 - p_y^*(\mathbf{y}))}{\|\mathbf{y} - \mathbf{p}^*(\mathbf{y})\|_2^3} &
\frac{(y_1 - p_x^*(\mathbf{y}))^2}{\|\mathbf{y} - \mathbf{p}^*(\mathbf{y})\|_2^3}
\end{bmatrix}
\end{equation}
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:2nd-deriv:H-Jgrad}
\mathbf{H}_{\mathcal{D}}(\mathbf{y}) =
\mathbf{J}_{\bm{\nabla}\mathcal{D}(\mathbf{y})}^T
\end{equation}

In practice, $\mathbf{H}_{\mathcal{D}}$ is computed numerically as the centered
finite difference of the normalized neighbor gradients
($\hat{\bm{\nabla}}\mathcal{D}$), as shown in
\eqref{eq:distance-maps:alignment:dmaps:2nd-deriv:H:numerical},
where $(\cdot)_{y_j}$ denotes the $j$-th component of the enclosed vector.
This numerical approximation is the discretization implemented in the proposed
approach for distance map-based 2D~point cloud registration
when requiring 2nd-order derivatives.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:2nd-deriv:H:numerical}
\mathbf{H}_{\mathcal{D}}(\mathbf{y}) \approx
\frac{1}{2\delta}
\begin{bmatrix}
\left(
    \hat{\bm{\nabla}}\mathcal{D}(y_1{+}\delta,\,y_2) -
    \hat{\bm{\nabla}}\mathcal{D}(y_1{-}\delta,\,y_2)
\right)_{y_1} &
\left(
    \hat{\bm{\nabla}}\mathcal{D}(y_1,\,y_2{+}\delta) -
    \hat{\bm{\nabla}}\mathcal{D}(y_1,\,y_2{-}\delta)
\right)_{y_1} \\[8pt]
\left(
    \hat{\bm{\nabla}}\mathcal{D}(y_1{+}\delta,\,y_2) -
    \hat{\bm{\nabla}}\mathcal{D}(y_1{-}\delta,\,y_2)
\right)_{y_2} &
\left(
    \hat{\bm{\nabla}}\mathcal{D}(y_1,\,y_2{+}\delta) -
    \hat{\bm{\nabla}}\mathcal{D}(y_1,\,y_2{-}\delta)
\right)_{y_2}
\end{bmatrix}
\end{equation}

\subsubsection{Representation levels}
\label{sec:distance-maps:alignment:dmaps:representation}

As for the representation of the distance map required for the proposed
registration process, each cell of the discrete grid may store different amounts
of precomputed information about the distance field. Therefore, we consider
three representation levels of increasing information and memory cost.
The choice of the representation determines which quantities are available at
$O(1)$ query cost during registration, and which must be recomputed on the fly
from neighboring cells or analytical expressions.

\paragraph{Level~0 -- $\mathcal{M}_0$ (\gls{nn} index only)}
Each cell stores only the integer index $k^*(\mathbf{y})$ of the
nearest reference point in $\mathcal{P}$
(see \eqref{eq:distance-maps:alignment:dmaps:representations:level0}).
The scalar distance $\mathcal{D}(\mathbf{y})$, the gradient vector
$\bm{\nabla}\mathcal{D}(\mathbf{y})$, and the Hessian matrix
$\mathbf{H}_{\mathcal{D}}(\mathbf{y})$ are not precomputed nor stored explicitly.
These values may be recovered at query time through their analytical expressions
discussed in \secref{sec:distance-maps:alignment:dmaps:derivation}.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:representations:level0}
\mathcal{M}_0(\mathbf{y}) = k^*(\mathbf{y})
= \argmin_{i \in \{1,\dots,n\}}
\|\mathbf{y} - \mathbf{p}_i\|_2 \in \mathbb{Z}
\end{equation}

Furthermore, $\mathcal{M}_0$ provides the most memory-efficient data
structure for data association, and may serve as a drop-in replacement
for the kd-tree \gls{nn} search~\cite{bentley:cacm:1975}
used in classical \gls{icp} methods~\cite{besl-mckay:tpami:1992,
chen-medioni:icra:1991}. This modification reduces the per-query cost from
$O(\log n)$ to $O(1)$ at the expense of
the precomputation step of $\mathcal{M}_0$.

\paragraph{Level~1 -- $\mathcal{M}_1$ (distance and first-order derivatives)}
Each cell stores the scalar distance $\mathcal{D}(\mathbf{y})$ together
with the two components of its spatial gradient
$\bm{\nabla}\mathcal{D}(\mathbf{y}) \in \mathbb{R}^2$, as shown in
\eqref{eq:distance-maps:alignment:dmaps:representations:level1}.
The integer index $k^*(\mathbf{y})$ of the nearest reference point in
$\mathcal{P}$
(see \eqref{eq:distance-maps:alignment:dmaps:representations:level0})
is optional for $\mathcal{M}_1$.
Hence, $\mathcal{M}_1$ requires three or four scalar values
per cell, depending on whether the $k^*(\mathbf{y})$ index is included into
$\mathcal{M}_1(\mathbf{y})$.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:representations:level1}
\mathcal{M}_1(\mathbf{y}) =
\left(
    \mathcal{D}(\mathbf{y}),
    \frac{\partial \mathcal{D}}{\partial y_1}(\mathbf{y}),
    \frac{\partial \mathcal{D}}{\partial y_2}(\mathbf{y})
\right)
\end{equation}

Subsequently, the scalar distance $\mathcal{D}(\mathbf{y})$ and
the gradient vector $\bm{\nabla}\mathcal{D}(\mathbf{y})$ are precomputed
based on \eqref{eq:distance-maps:alignment:dmaps:definition} and using the
central finite difference scheme of
\eqref{eq:distance-maps:alignment:dmaps:1st-deriv:numerical-derivation},
respectively.
The Hessian matrix $\mathbf{H}_{\mathcal{D}}(\mathbf{y})$ is not precomputed nor
stored explicitly. This value may be recovered at query time through its
analytical expression or through the central finite difference approximation
from neighbor cells of the normalized gradient discussed in
\secref{sec:distance-maps:alignment:dmaps:derivation}.

\paragraph{Level~2 -- $\mathcal{M}_2$
(distance, first-, and second-order derivatives)}
Although the analysis in \secref{sec:distance-maps:alignment:dmaps:derivation}
demonstrates that the analytical expression for $\mathbf{H}_{\mathcal{D}}$
leads to a symmetric matrix, its centered finite difference approximation
presented in \eqref{eq:distance-maps:alignment:dmaps:2nd-deriv:H:numerical} does
not ensure that property.
Therefore, each cell of $\mathcal{M}_2$ additionally stores
all entries of the approximated Hessian matrix
$\mathbf{H}_{\mathcal{D}}(\mathbf{y})\in\mathbb{R}^{2\times2}$, as shown in
\eqref{eq:distance-maps:alignment:dmaps:representations:level2}.
Similarly to $\mathcal{M}_1$, the integer index $k^*(\mathbf{y})$
(see \eqref{eq:distance-maps:alignment:dmaps:representations:level0})
is optional for the $\mathcal{M}_2$ representation level.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:representations:level2}
\mathcal{M}_2(\mathbf{y}) =
\left(
    \mathcal{D}(\mathbf{y}),
    \frac{\partial \mathcal{D}}{\partial y_1}(\mathbf{y}),
    \frac{\partial \mathcal{D}}{\partial y_2}(\mathbf{y}),
    \frac{\partial^2 \mathcal{D}}{\partial y_1^2}(\mathbf{y}),
    \frac{\partial^2 \mathcal{D}}{\partial y_1 \partial y_2}(\mathbf{y}),
    \frac{\partial^2 \mathcal{D}}{\partial y_2 \partial y_1}(\mathbf{y}),
    \frac{\partial^2 \mathcal{D}}{\partial y_2^2}(\mathbf{y})
\right)
\end{equation}

The representation level $\mathcal{M}_2$ makes available all values throughout
the registration process, thanks to the precomputation of the scalar distance
$\mathcal{D}(\mathbf{y})$ stored in the discrete grid of the distance and based
on \eqref{eq:distance-maps:alignment:dmaps:definition}, while both the
gradient vector $\bm{\nabla}\mathcal{D}(\mathbf{y})$ and the Hessian matrix
$\mathbf{H}_{\mathcal{D}}(\mathbf{y})$ are computed by centered finite
differences from neighbor cells based on
\eqref{eq:distance-maps:alignment:dmaps:1st-deriv:numerical-derivation} and
\eqref{eq:distance-maps:alignment:dmaps:2nd-deriv:H:numerical}, respectively.

\subsubsection{Computation and data structures}
\label{sec:distance-maps:alignment:dmaps:computation}
Independently of the representation level, the computation of $\mathcal{M}_l$
is based on a wavefront propagation algorithm using a best-first priority queue,
analogous to Dijkstra's shortest-path algorithm~\cite{dijkstra:nm:1959}.
All variants considered in our approach share the same high-level pipeline:
\begin{enumerate}[nosep]
\item initialize the grid based on its resolution $\delta$ (m/px),
      the bounding box of $\mathcal{P}$, optional fixed map size (m), and
      optional truncation distance $d_{\max}$ (m);
\item seed the priority queue with the grid cells occupied by reference
      points, initializing their distance to zero;
\item propagate the integer index $k^*(\mathbf{y})$ of the nearest
      reference point and its corresponding distance value outward via
      best-first wavefront expansion through the $\mathcal{N}_8$ neighborhood
      (8-connectivity) of each cell in the priority queue for the whole map or
      until $d_{\max}$ is achieved from the nearest reference point;
\item depending on the required representation level
      $\mathcal{M}_l$, execute post-processing passes to compute the
      gradient $\bm{\nabla}\mathcal{D}$ and the Hessian
      $\mathbf{H}_{\mathcal{D}}$ through centered finite differences
      (see \secref{sec:distance-maps:alignment:dmaps:derivation}).
\end{enumerate}

\paragraph{Coordinates conversion}
All implemented variants in our approach for distance map-based representations
use the same affine mapping between metric coordinates
$\mathbf{y}\in\mathbb{R}^2$ \gls{wrt} the sensor coordinate frame and
integer grid indices $(r,c)\in\mathbb{Z}^2$,
and vice-versa. Given $\mathbf{y}_{\mathcal{M},\min}\in\mathbb{R}^2$,
the metric coordinates corresponding to the cell
$(r,c)_{\mathcal{M},\min}=(0,0)$ of the grid, determined during
the initialization procedure of the representation level $\mathcal{M}_l$,
\eqref{eq:distance-maps:alignment:dmaps:conversion:global2local} defines
the mapping from the sensor coordinates to the grid indices.
Note that negative indices or values that exceed the grid size
represent invalid cells outside the data structure.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:conversion:global2local}
(r, c) =
\left\lfloor\frac{\mathbf{y} - \mathbf{y}_{\mathcal{M},\min}}{\delta}\right\rfloor
\in\mathbb{Z}^2
\end{equation}

Furthermore,
\eqref{eq:distance-maps:alignment:dmaps:conversion:global2local:real}
formulates a sub-pixel (non-integer) version of
\eqref{eq:distance-maps:alignment:dmaps:conversion:global2local}, omitting
the floor operator. We use this version for bilinear interpolation at
query time, enabling sub-pixel accuracy for the distance and derivatives
values returned to the registration optimizer.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:conversion:global2local:real}
(r, c)_{\mathbb{R}} =
\frac{\mathbf{y} - \mathbf{y}_{\mathcal{M},\min}}{\delta}
\in \mathbb{R}
\end{equation}

Lastly, the inverse mapping from grid indices to metric coordinates is obtained
by multiplying the indices by the resolution $\delta$ and offsetting by
$\mathbf{y}_{\mathcal{D},\min}$, as formulated in
\eqref{eq:distance-maps:alignment:dmaps:conversion:local2global}.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:conversion:local2global}
\mathbf{y} = \delta \cdot (r, c)^T + \mathbf{y}_{\mathcal{M},\min} \in \mathbb{R}^2
\end{equation}

\paragraph{Data structure storage}
Independently of the required
$\mathcal{M}_l$, we store the discrete grid in one of two memory layouts.
Under \emph{dense storage} ($\mathcal{M}_{l,\mathrm{dense}}$), the grid is
allocated as a contiguous array of cells (generic
Eigen\footnote{\label{note:eigen}\url{https://libeigen.gitlab.io/}} matrix in
C++), where every cell in the grid is populated during precomputation.
Under \emph{sparse storage}, only cells satisfying
$\mathcal{D}(\mathbf{y})\leq d_{\max}$ are stored, since moving points
projecting further than $d_{\max}$ distance from any reference point
are excluded from the registration cost. This truncated representation is
analogous to the \glslink{tsdf}{\glsentryfull{tsdf}} used in volumetric
reconstruction~\cite{curless-levoy:siggraph:1996},
where only a narrow band of the distance values is retained around a surface.

Moreover, we implement the sparse grid in C++ as a hash table using the container
\texttt{std::unordered\_map}\footnote{\label{note:hash-table}\url{https://en.cppreference.com/w/cpp/container/unordered_map.html}}.
The hash key for a cell $(r,c)$ is computed by encoding both integer indices
(\texttt{int32}) into a single 64-bit unsigned integer, as formulated in
\eqref{eq:distance-maps:alignment:dmaps:computation:hash-key}, where
$\ll$ denotes a bit-wise left shift and $|$ denotes a bit-wise OR.
An important property of \texttt{std::unordered\_map} is that the container does
not invalidate existing element pointers upon rehashing, allowing the
priority queue to store raw cell pointers safely throughout the entire
wavefront propagation. After propagation, a rehash call consolidates
the bucket layout before the gradient and Hessian post-processing passes
iterate over all stored entries.
\begin{equation}
\label{eq:distance-maps:alignment:dmaps:computation:hash-key}
\mathrm{key}(r, c) =
\bigl(\texttt{uint64}(r) \ll 32\bigr)
\;\big|\;
\bigl(\texttt{uint64}(c) \;\&\; \texttt{0xFFFFFFFF}\bigr)
\end{equation}

\paragraph{Computation of the data structures}

First, \algref{alg:distance-maps:alignment:dmaps:initialization} details the
initialization procedure for the grid data structure.
This structure is parametrized by a map size ($L$) and a truncation distance
($d_{\max}$). If $L > 0$, the metric bounding box \gls{wrt} the local coordinate
frame of the sensor for the distance map is symmetrically defined as
$[-L, L]$. Otherwise, the bounding box is computed directly from
the reference point cloud $\mathcal{P}$ by iterating over all valid points and
accumulating the minimum and maximum coordinates.
Then, the grid dimensions in pixels are derived from the bounding box.
If $d_{\max} > 0$ is set and no explicit map size was
provided, the bounding box is expanded by $d_{\max}$ on both sides before
converting to pixel cells, ensuring that the truncation band is fully covered
around the point cloud boundary.
Otherwise, the bounding box is converted to pixel cell counts using
the resolution $\delta$ based on the original metric bounding box limits.
Finally, the grid is resized to the computed dimensions. The subsequent
initialization of cell values depends on the underlying data structure, as
detailed in \algref{alg:distance-maps:alignment:dmaps:nano},
\algref{alg:distance-maps:alignment:dmaps:dense}, and
\algref{alg:distance-maps:alignment:dmaps:sparse}.

\input{algorithms/distance-maps/distance-map_init.tex}

Next, \algref{alg:distance-maps:alignment:dmaps:nano},
\algref{alg:distance-maps:alignment:dmaps:dense}, and
\algref{alg:distance-maps:alignment:dmaps:sparse} present the computation
procedure for each type of distance map structures considered in the scope of
this PhD~Thesis: \emph{Nano} ($\mathcal{M}_0$, dense storage), \emph{Standard}
($\mathcal{M}_1$ or $\mathcal{M}_2$, dense storage), and \emph{Sparse}
($\mathcal{M}_1$ or $\mathcal{M}_2$, sparse storage), respectively.
Although the $\mathcal{M}_0$ representation could be implemented using
a sparse storage layout, the overhead of \texttt{std::unordered\_map},
which may reserve significantly more hash table buckets than
the number of elements actually stored, does not justify a sparse structure for
a single integer index per cell.

\input{algorithms/distance-maps/distance-map_nano.tex}
\input{algorithms/distance-maps/distance-map_standard.tex}
\input{algorithms/distance-maps/distance-map_sparse.tex}

Furthermore, \algref{alg:distance-maps:alignment:dmaps:nano} details
the full construction of the \emph{Nano} variant.
First, the entire grid $\mathcal{M}_0$ is filled with $-1$, representing
unvisited cells, and the priority queue $\mathcal{Q}$ is cleared.
Next, the reference point cloud $\mathcal{P}$ is projected onto the grid
using the coordinate conversion $\mathrm{global2local}(\cdot)$ of
\eqref{eq:distance-maps:alignment:dmaps:conversion:global2local}.
Invalid points and points whose projected cell $(r, c)$ lies outside
the grid are skipped. For each remaining point $\mathbf{p}_i$,
the corresponding cell is assigned the index $i$.
A pointer to the cell is then pushed into $\mathcal{Q}$ with distance
$d = 0$, which sets the priority of the queue entry.
Since $\mathcal{M}_0$ does not store the distance value, the queue
entry must carry it explicitly alongside the cell pointer, so that
$\mathcal{Q}$ correctly orders cells during wavefront propagation.

At each iteration of the wavefront propagation, the element $q_k$ with
the smallest distance is popped from $\mathcal{Q}$. A negative stored \gls{nn}
index ($k < 0$) indicates an entry overwritten by a closer parent that surpasses
the truncation distance $d_{\max}$ when set ($d_{\max} > 0$), and is therefore
discarded. For each cell $(r', c')$ in the $\mathcal{N}_8(r_k, c_k)$
neighborhood, the neighbor cell is updated to the closer parent provided
the truncation condition is satisfied, \ie,
\ $d_k < d_{\max}$ when truncation is enabled, or
unconditionally true when the full map is computed.
The first case is when the cell is unvisited ($o < 0$). The other case is
if already visited ($o \geq 0$) but with a stored distance greater than
the candidate ($d_o > d_k$). In both cases, the updated cell is pushed into
$\mathcal{Q}$, allowing the wavefront propagation to correct sub-optimal
assignments from earlier iterations.

The \emph{Standard} variant detailed in
\algref{alg:distance-maps:alignment:dmaps:dense} follows a similar construction
pipeline as the \emph{Nano} variant
(see \algref{alg:distance-maps:alignment:dmaps:nano}), with two key
differences arising from the more complete cell data structure.
Since the distance is now part of the cell itself in $\mathcal{M}_1$ and
$\mathcal{M}_2$ representation levels, the priority queue $\mathcal{Q}$
does not need to carry the distance separately, accessing directly from the
representation of the distance map.
The other difference is the distinction between unvisited and already-visited
neighbor cells, which required an explicit branch in
\algref{alg:distance-maps:alignment:dmaps:nano}, is no longer necessary.
Since all cells are initialized with $+\inf$, the single condition
$\left(d_k < d_{\max} \textbf{ or } d_{\max} \leq 0\right) \textbf{ and }d_o > d_k$
handles both cases uniformly, where an unvisited cell holds a $+\inf$ distance,
while a previously visited cell is updated only if a strictly closer parent
is found.
After the wavefront terminates, the gradient
$\bm{\nabla}\mathcal{D}$ is computed over the entire dense grid via the
central finite difference scheme of
\eqref{eq:distance-maps:alignment:dmaps:1st-deriv:numerical-derivation}.
If the $\mathcal{M}_2$ representation is requested, a second
post-processing pass computes the Hessian $\mathbf{H}_{\mathcal{D}}$
via \eqref{eq:distance-maps:alignment:dmaps:2nd-deriv:H:numerical}.
Both passes use one-sided (forward or backward) differences at grid borders.

As for the distance computation on the \emph{Sparse} variant described in
\algref{alg:distance-maps:alignment:dmaps:sparse}, its construction differs
from the \emph{Standard} variant
(see \algref{alg:distance-maps:alignment:dmaps:dense}) in three steps arising
from the \texttt{std::unordered\_map} storage layout. The
\texttt{std::unordered\_map} is not pre-allocated for the entire map, in order
to maintain a sparse storage layout when the truncation distance is set
($d_{\max} > 0$). Instead, the implemented wrapper around
\texttt{std::unordered\_map} defines a default cell whenever a queried index
$(r, c)$ is not yet present in the hash table.
Next, each entry pushed into the priority queue $\mathcal{Q}$ carries more
information than in the \emph{Standard} variant: a pointer to the
cell $\mathcal{M}_{1,2}(r, c)$, the current cell index
$\{(r, c)\}$, and the parent \gls{nn} index $\{(r_p, c_p)\}$.
This implementation avoids additional hash queries at every propagation step.
Lastly, once the wavefront terminates and before precomputing the derivatives,
a rehash call is issued to consolidate the internal bucket layout of the
\texttt{std::unordered\_map} to match the final number of populated
cells, minimizing the load factor and reducing hash collision overhead
for the subsequent read-intensive derivative passes.

Table~\ref{tab:distance-maps:alignment:dmaps} summarizes
the three distance map variants considered in this PhD~Thesis in
terms of storage layout, memory occupation, query lookup complexity, and
truncation requirements.
The \emph{Nano} variant stores a single integer index per cell, resulting
in the lowest memory footprint of $O(N_G)$.
Both the \emph{Standard} and \emph{Sparse} variants store between $3$
and $8$ scalar values per cell depending on the representation level
$\mathcal{M}_{1,2}$ and whether the optional \gls{nn} index
$k^*(\mathbf{y})$ is included.
While the \emph{Standard} variant allocates the full grid as a dense
array with $O(N_G)$ memory, the \emph{Sparse} variant backed by
\texttt{std::unordered\_map} populates only the $N_d \leq N_G$ cells
within the truncation band, making it the most memory-efficient choice
when $d_{\max}$ is set.
All three variants provide $O(1)$ query lookup, though the \emph{Sparse}
variant achieves this only on average, with worst-case $O(N_d)$ under
hash collision, mitigated by the post-propagation rehash discussed previously.

\input{tables/distance-maps/variants.tex}

\subsection{Problem formulation}
\label{sec:distance-maps:alignment:problem}

Let $\mathbf{z}_i\in\mathbb{R}^2$, $i = 1, \dots, M$, represent
the 2D~point measurements from the moving (current) point cloud.
Similarly to the reference cloud $\mathcal{P}$, the moving cloud is also
acquired by a 2D~laser scanner and expressed in the sensor coordinate frame.
The goal of 2D~point cloud registration is to find the robot pose
$\mathbf{x} \in SE(2)$ that best aligns the moving cloud to the reference cloud.
The robot pose is represented as a homogeneous transformation on the $SE(2)$
manifold, as formulates in \eqref{eq:distance-maps:alignment:problem:x:pose},
where $\mathbf{R}(\theta) \in SO(2)$ is the 2D orthonormal rotation matrix
(see \eqref{eq:distance-maps:alignment:problem:x:rotation}) and
$\mathbf{t} = (t_x, t_y)^T \in \mathbb{R}^2$ is the translation vector.
\begin{equation}
\label{eq:distance-maps:alignment:problem:x:pose}
\mathbf{x} = \left[\mathbf{R} \mid \mathbf{t}\right] \in SE(2)
\end{equation}
\begin{equation}
\label{eq:distance-maps:alignment:problem:x:rotation}
\mathbf{R}(\theta) =
\begin{bmatrix} r_{1,1} & r_{1,2} \\ r_{2,1} & r_{2,2} \end{bmatrix} =
\begin{bmatrix} \cos(\theta) & -\sin(\theta) \\ \sin(\theta) & \cos(\theta) \end{bmatrix} \in SO(2)
\end{equation}

Furthermore, let $\mathbf{T}_S^R \in SE(2)$ denote the known rigid-body
transformation of the sensor \gls{wrt} the robot frame (the extrinsic
calibration of the laser scanner), as formulated in
\eqref{eq:distance-maps:alignment:problem:sensor}. Then,
\eqref{eq:distance-maps:alignment:problem:sensor:inv} formulates the inverse
transformation for the extrinsic parameters of the sensor ($\mathbf{T}_R^S$).
As a result, the moving point $\mathbf{z}_i$ expressed in the robot frame is
formulated by \eqref{eq:distance-maps:alignment:problem:moving-pt:robot}, while
the same point transformed by the current pose estimate $\mathbf{x}$ is defined
by \eqref{eq:distance-maps:alignment:problem:moving-pt:projected}.
\begin{equation}
\label{eq:distance-maps:alignment:problem:sensor}
\mathbf{T}_S^R = \left[\mathbf{R}_S^R \mid \mathbf{t}_S^R\right] \in SE(2)
\end{equation}
\begin{equation}
\label{eq:distance-maps:alignment:problem:sensor:inv}
\mathbf{T}_R^S =
\left[\mathbf{R}_R^S \mid \mathbf{t}_R^S\right] =
\left[{\mathbf{R}_S^R}^T \mid -{\mathbf{R}_S^R}^T\mathbf{t}_S^R\right]
\in SE(2)
\end{equation}
\begin{equation}
\label{eq:distance-maps:alignment:problem:moving-pt:robot}
\mathbf{z}_{R,i} = \mathbf{T}_S^R \mathbf{z}_i
\end{equation}
\begin{equation}
\label{eq:distance-maps:alignment:problem:moving-pt:projected}
\tilde{\mathbf{z}}_{R,i} = \mathbf{x}_R \mathbf{T}_S^R \mathbf{z}_i
\end{equation}

Since the optimization variable $\mathbf{x}$ lives on the $SE(2)$ manifold,
incremental updates cannot be performed by simple Euclidean addition,
as discussed in \secref{sec:distance-maps:background:least-squares:manifold}.
A perturbation
$\Delta\mathbf{x} = (\Delta x, \Delta y, \Delta\theta)^T \in \mathbb{R}^3$ is
defined in the tangent (Euclidean) space with the exponential and logarithmic
mappings defined as
\eqref{eq:distance-maps:alignment:problem:x:exp} and
\eqref{eq:distance-maps:alignment:problem:x:log}, respectively.
Consequently, the manifold update between Gauss-Newton iterations adopt a
right-sided perturbation update defined in
\eqref{eq:distance-maps:alignment:problem:gauss-newton:addition}, with the
corresponding inverse (distance) operation as formulated by
\eqref{eq:distance-maps:alignment:problem:gauss-newton:subtraction}.
\begin{equation}
\label{eq:distance-maps:alignment:problem:x:exp}
\exp\!\left(\Delta\mathbf{x}\right) =
\left[\mathbf{R}(\Delta\theta) \mid (\Delta x, \Delta y)^T \right] \in SE(2)
\end{equation}
\begin{equation}
\label{eq:distance-maps:alignment:problem:x:log}
\log(\mathbf{x})_{\mathbf{x} \in SE(2)} =
    \left(t_x,t_y,\atan2\!\left(r_{2,1},\, r_{1,1}\right)\right)^T
    \in \mathbb{R}^3
\end{equation}
\begin{equation}
\label{eq:distance-maps:alignment:problem:gauss-newton:addition}
\mathbf{x}_{k+1} = \mathbf{x}_k \oplus \Delta\mathbf{x}
                 = \mathbf{x}_k \cdot \exp\!\left(\Delta\mathbf{x}\right)
\end{equation}
\begin{equation}
\label{eq:distance-maps:alignment:problem:gauss-newton:subtraction}
\Delta\mathbf{x} = \mathbf{x}_1 \ominus \mathbf{x}_2
                 = \log\!\left(\mathbf{x}_1^{-1}\, \mathbf{x}_2\right)^\vee
\end{equation}

Lastly, the cross-product matrix (hat operator) in 3D maps a vector
$\mathbf{v} \in \mathbb{R}^3$ to a $3\times 3$ skew-symmetric matrix and gives
the rotational sensitivity of a rotated vector. In 2D, an analogous operator
is defined for a vector $\mathbf{v} = (v_x, v_y)^T \in \mathbb{R}^2$, as shown
in \eqref{eq:distance-maps:alignment:problem:skew}.
\begin{equation}
\label{eq:distance-maps:alignment:problem:skew}
\lfloor \mathbf{v} \rfloor =
\left.\frac{\partial\mathbf{R}(\theta)\mathbf{v}}{\partial \theta}
\right|_{\theta=0} =
\begin{bmatrix} 0 & -1 \\ 1 & \phantom{-}0 \end{bmatrix} \cdot \mathbf{v} =
\begin{pmatrix} -v_y \\ v_x \end{pmatrix}
\end{equation}

Intuitively, $\lfloor \mathbf{v} \rfloor$ is the 90\textdegree{}
counter-clockwise rotation of $\mathbf{v}$, \ie\ the instantaneous velocity of
$\mathbf{v}$ under an infinitesimal positive rotation. This skew operator
appears in all Jacobian matrices of the 2D~point cloud registration problem
derived in \secref{sec:distance-maps:alignment}, corresponding to the column
associated with the rotational degree of freedom $\Delta\theta$ of
the perturbation $\Delta\mathbf{x}$.

\subsection{Point-to-point error formulation}
\label{sec:distance-maps:alignment:pt2pt}

\subsubsection{Jacobian matrix via right-sided perturbation}

\subsection{Point-to-plane error formulation}
\label{sec:distance-maps:alignment:pt2pl}

\subsubsection{Jacobian matrix via right-sided perturbation}
\subsubsection{Normals estimation}

\section{Distance Maps}\label{sec:distance-maps}

TBC (unsigned distance map structure,
types of distance map formulations -- only NN, NN + distance,
NN + 1st derivative, NN + 1st and 2nd derivative).

\section{Data Registration}\label{sec:distance-maps:data-registration}

TBC

\subsection{Least-squares on manifolds}

TBC

\subsection{Optimization algorithm}
\subsection{Covariance estimation}

\subsection{Problem formulation -- point-to-point error formulation}

TBC

\subsubsection{State space}
\subsubsection{Measurement space}
\subsubsection{Prediction function}
\subsubsection{Error function}

\subsection{Problem formulation -- point-to-plane error formulation}

TBC

\subsubsection{State space}
\subsubsection{Measurement space}
\subsubsection{Prediction function}
\subsubsection{Error function}

\section{Experimental Results and Discussion}\label{sec:distance-maps:results}

TBC

\subsection{Alignment results}
\subsection{Covariance estimation}
\subsection{Computation performance}

\section{Summary and Conclusions}\label{sec:distance-maps:conclusions}

TBC
