\chapter{Line Fitting-Based Corner-Like Detector for 2D Laser Scanners Data}\label{chap:features}

TBC

\section{Background}\label{sec:features:background}

TBC (background / literature analysis specific to the topic of this chapter)

\subsection{Line fitting intersection}\label{sec:features:background:line-fitting}

A trend in the literature is to extract corners at the intersections of
line segments detected in the 2D~laser
scan~\cite{castellanos:mfi:1996,roumeliotis-bekey:iros:2000,jung:tits:2008,falomir:prl:2013,fernandez-moral:ijrr:2015,mohamed:isprs-arxiv:2016,zhao:icsess:2016,lin:yac:2017,qu:robio:2018,liu:electronics:2022,liu:sensors:2023}.
First, the laser data may be filtered to remove invalid values
(zero range values~\cite{jung:tits:2008}) and possible noise
(\eg, considering the minimum distance to the two neighbors of
a certain point to remove isolated data~\cite{jung:tits:2008},
mean average filtering~\cite{liu:electronics:2022}, or bilateral filters
to remove non-linear noise~\cite{liu:sensors:2023}).
Next, a segmentation step extracts sets of continuous points
in the environment for line fitting estimation.
Castellanos~\etal~\cite{castellanos:mfi:1996} segment the scan into
polygonal lines and subsequently divide them into
segments using an iterative method.
Falomir~\etal~\cite{falomir:prl:2013} search for discontinuities
in the data to form point subsets and use a recursive line-fitting
algorithm~\cite{xavier:icra:2005} to obtain line segments.
Fernández-Moral~\etal~\cite{fernandez-moral:ijrr:2015} search for
parameters of a 2D~line that maximize the number of points
supporting the line model using the \gls{ransac}
algorithm~\cite{fischler-bolles:cacm:1981}.
Mohamed~\etal~\cite{mohamed:isprs-arxiv:2016} use the line tracking
algorithm~\cite{munoz:ropec:2013} to iteratively segment lines
in the scan and perform \gls{pca} to compute the lines' parameters.
Then, the intersection points between consecutive line segments
define a corner feature in the environment.

Furthermore, Lin~\etal~\cite{lin:yac:2017} and
Liu~\etal~\cite{liu:electronics:2022,liu:sensors:2023} use adaptive threshold
segmentation to cluster range data into continuous spatial subsets.
Although Lin~\etal~\cite{lin:yac:2017} state that an adaptive segmentation
was used due to the non-uniform distribution of the 2D~laser data,
the authors do not specify the threshold used in their work.
In Liu~\etal~\cite{liu:electronics:2022} work, the threshold is based on
the laser's angular resolution and the distance to
the current point under analysis.
Liu~\etal~\cite{liu:sensors:2023} also establish a threshold depending on
the angular resolution. Instead of considering only the distance of
the point under analysis, the proposed approach~\cite{liu:sensors:2023} selects
the larger distance between adjacent points.
Even though Roumeliotis and Bekey~\cite{roumeliotis-bekey:iros:2000} do not use
a specific adaptive threshold, the proposed SEGMENTS algorithm with two-level
\gls{ekf} exhibits adaptive behavior. A stricter filter,
in terms of the system's noise covariance, attempts to follow a straight line,
while a more flexible \gls{ekf} tries to follow a new segment during each scan.
The four methods~\cite{lin:yac:2017,liu:electronics:2022,liu:sensors:2023,roumeliotis-bekey:iros:2000}
also define a corner feature as the intersection
between consecutive line segments.

A specific application of corner detection with 2D~laser data in the literature
is vehicle detection~\cite{jung:tits:2008,qu:robio:2018}.
Jung~\etal~\cite{jung:tits:2008} cluster laser data using a fixed threshold
between consecutive range values. Then, the method performs template matching
for every point under analysis, assuming the point is
the vertex of the L-shape of a vehicle in the scan.
Jung~\etal~\cite{jung:tits:2008} use \gls{svd} to estimate the L-shape's
line parameters and decompose the problem into line-curve estimation
when the vehicle has a curved corner in the data.
In contrast, Qu~\etal~\cite{qu:robio:2018} segment the data using
\gls{dbscan}~\cite{ester:kdd:1996}.
The method performs L-shape fitting in two steps:
searching for the L-shape vertices and corner-point localization,
the latter based on \gls{ransac}~\cite{fischler-bolles:cacm:1981}.

However, corner detection algorithms that rely on global search during
line segment scanning may be limited to well-structured environments.
Thus, we propose a corner-like detection method based on local line fitting,
without requiring global line detection or recursive data search.
Also, the experiments presented in \secref{sec:features:results} demonstrate
the ability of the proposed detector to extract more points of interest
from the scan in addition to environment corners.

\subsection{Curvature-based detection}\label{sec:features:background:curvature}

Other approaches to detecting corners in 2D~laser data estimate the curvature.
N\'{u}\~{n}ez~\etal~\cite{nunez:icra:2006} propose an adaptive curvature
estimator that compares the $k$-nearest points on
both sides of the point being evaluated.
The $k$ size is adjusted based on the distance between possible corners.
A segmentation step precedes the curvature estimation to detect breakpoints
using the adaptive breakpoint detector proposed by
Borges~and~Aldon~\cite{borges-aldon:jint:2004}.
Then, N\'{u}\~{n}ez~\etal~\cite{nunez:icra:2006} extract corner, line, and
curve segment features from the scan data by analyzing whether
there are peaks in curvature, flat segments with zero average curvature,
and flat segments with non-zero average curvature, respectively.
V\'{a}zquez-Mart\'{i}n~\etal~\cite{vazquez-martin:sensors:2009} use
the adaptive curvature estimator of N\'{u}\~{n}ez~\etal~\cite{nunez:icra:2006}
to segment the scan into clusters of homogeneous curvature.
Even though the existence of a corner is determined from the curvature function,
its keypoint definition is based on the intersection of the two line features
that generate the corner, to improve robustness to noise.
Also, Certad~\etal~\cite{certad:andescon:2012} modify the work of
N\'{u}\~{n}ez~\etal~\cite{nunez:icra:2006} to consider
two curvature estimations based on different noise-related thresholds,
thereby improving the rate of true detection for
corner and edge features in the experiments.

Moreover, Feng~\etal~\cite{feng:aici:2009} and
Zhao~\etal~\cite{zhao:icsess:2016} report a correlation between
the angle difference between the two fitted lines on
both sides of a scan point and the curvature value.
Feng~\etal~\cite{feng:aici:2009} introduce the micro-tangent line function
as the variance between the tangents at two consecutive points in the scan.
Similarly to N\'{u}\~{n}ez~\etal~\cite{nunez:icra:2006},
line and curve segments are extracted from consecutive range readings
with variance close to zero and with constant value larger than a threshold,
respectively, and corner features are extracted from peaks in the variance
function. Instead of only considering two consecutive points,
Zhao~\etal~\cite{zhao:icsess:2016} define two distance-based regions on
both sides of a point to estimate direction vectors and,
consequently, the angle difference between those vectors.
Candidate corner points are selected based on a fixed threshold
for the angle difference. Then, the corner point is estimated
via non-minimal suppression.

Still, most curvature-based methods for corner feature extraction associate
the keypoint with a laser reading. Given the discrete, radial nature
of 2D~laser, the estimated corner may vary farther away.
Also, there is the question of whether the curvature estimators
are invariant to scale, angular resolution, and different viewpoints.
We propose a detector based on the stability of intersection points
from fitted lines at different distance-based scales.
The experiments in \secref{sec:features:results} focus
on evaluating the invariance of the proposed corner-like features.

\section{Corner-Like Feature Detector}

TBC

\subsection{Linear least-squares regression}
\subsection{Feature extraction}

\section{Experimental Results and Discussion}\label{sec:features:results}

TBC

\subsection{Feature extraction results}
\subsection{Features observability}
\subsection{Features association distance}
\subsection{Integration with a line features extractor algorithm}

\section{Summary and Conclusions}\label{sec:features:conclusions}

TBC
